---
title: "MasterDoc"
author: "Nora Otic"
date: "2025-06-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

# Group 4 Data Analysis Project 

# Silent Signals: 
# Projekt Overview

## Objective
The aim of this project is the automated identification of RNA-dependent proteins in a non-synchronized A549 cell line, which originates from a lung adenocarcinoma.
To achieve this, we investigate the influence of RNA on protein complexes by comparing protein distribution profiles between untreated cell lysates (control samples) and RNase-treated lysates (RNase samples

## Steps
- 1. data exploration
- 2. data normalisation
- 3. shift analysis
- 4. dimension reduction
- 5. detection of features predictive of RNA-binding

## libraries
- limma
- pheatmap
- sva
- stringr
- emdist
- DT
- tidyr
- ggplot2
- dplyr
- shiny
- cluster
- factoextra
- scatterplot3d
- plotly

# load data
```{r}
#load dataset
url <- "https://hub.dkfz.de/s/QT3BfRperQqMxPF/download/RDeeP_A549_NS.csv.zip"


#Create a temporary file and download the ZIP
temp_file <- tempfile()
download.file(url, temp_file)
zip_contents <- unzip(temp_file, list = TRUE)
print(zip_contents)

csv_file <- unzip(temp_file, files = zip_contents$Name[1], exdir = tempdir())
A549_NS <- read.csv(csv_file, sep = ";")
print(A549_NS)
```

# 1. data exploration
```{r}
#data exploration
print(A549_NS)

#number of rows and columns
nrow(A549_NS)
ncol(A549_NS)

#type of data (überprüft die class der Spalten -> alle Eintrage haben zwangsweise dieselbe class)
sum(sapply(A549_NS, is.numeric))

#finding NAs
sum(is.na(A549_NS))
```

```{r}
# all columns
col_names <- colnames(A549_NS)

# identify rnase and control columns 
ctrl_cols <- c("Protein_Name", grep("Ctrl", col_names, value = TRUE))
rnase_cols <- c("Protein_Name", grep("RNase", col_names, value = TRUE))

# create dataframe
df_ctrl <- A549_NS[, ctrl_cols, drop = FALSE]
df_rnase <- A549_NS[, rnase_cols, drop = FALSE]

head(df_ctrl)
head(df_rnase)


```


# Reproducability
Goal: evaluate the consistency of the measured replicates across the fractions for both treatments to determine the reproducability 
## Reproducibility of ctrl values
```{r}
# Reproducabilty of control measurements 
# create dataframe for analysis of the replicates
ctrl_correlations <- data.frame(
  fraction = character(),
  ctrl1_vs_ctrl2 = numeric(),
  ctrl1_vs_ctrl3 = numeric(),
  ctrl2_vs_ctrl3 = numeric(),
  stringsAsFactors = FALSE
)

# loop over all fractions and calculate spearman correlation
for (i in 0:24) {
  start_col <- 2 + i * 3 
  end_col <- start_col + 2
  ctrl_data <- df_ctrl[, start_col:end_col]
  colnames(ctrl_data) <- c("ctrl1", "ctrl2", "ctrl3")
  cor_matrix <- cor(ctrl_data, use = "pairwise.complete.obs", method = "spearman")
  ctrl_correlations <- rbind(ctrl_correlations, data.frame(
    fraction = paste0("fraction", i + 1),
    ctrl1_vs_ctrl2 = cor_matrix["ctrl1", "ctrl2"],
    ctrl1_vs_ctrl3 = cor_matrix["ctrl1", "ctrl3"],
    ctrl2_vs_ctrl3 = cor_matrix["ctrl2", "ctrl3"]
  ))
}

# show result
print(ctrl_correlations)

library(ggplot2)
library(tidyr)

ctrl_cor_long <- pivot_longer(ctrl_correlations, 
                               cols = starts_with("ctrl"), 
                               names_to = "comparison", 
                               values_to = "correlation")

# Plot 
ggplot(ctrl_cor_long, aes(x = fraction, y = correlation, color = comparison)) +
  geom_point(size = 3) +
  scale_color_manual(values = c(
    "ctrl1_vs_ctrl2" = "#c9a0dc", 
    "ctrl1_vs_ctrl3" = "#A9C1A6",  
    "ctrl2_vs_ctrl3" = "#F5E1A4"   
  )) +
  ylim(0.4, 1) +  
  theme_minimal() +
  labs(title = "Spearman Correlation of Controls",
       y = "Spearman ρ",
       x = "Fraction",
       color = "Replicate Comparison") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5),
        plot.title = element_text(hjust = 0.5))
```
Output: very high correlation across the fractions shows reproducability for the control measurments
-> all above 0.8 


## Reproducibility of rnase values
```{r}
# Reproducabilty of rnase measurements 
# create dataframe for analysis of the replicates
rnase_correlations <- data.frame(
  fraction = character(),
  rnase1_vs_rnase2 = numeric(),
  rnase1_vs_rnase3 = numeric(),
  rnase2_vs_rnase3 = numeric(),
  stringsAsFactors = FALSE
)

# loop over all fractions and calculate the spearman correlation
for (i in 0:24) {
  start_col <- 2 + i * 3 
  end_col <- start_col + 2
  rnase_data <- df_rnase[, start_col:end_col]
  colnames(rnase_data) <- c("rnase1", "rnase2", "rnase3")
  cor_matrix <- cor(rnase_data, use = "pairwise.complete.obs", method = "spearman")
  rnase_correlations <- rbind(rnase_correlations, data.frame(
    fraction = paste0("fraction", i + 1),
    rnase1_vs_rnase2 = cor_matrix["rnase1", "rnase2"],
    rnase1_vs_rnase3 = cor_matrix["rnase1", "rnase3"],
    rnase2_vs_rnase3 = cor_matrix["rnase2", "rnase3"]
  ))
}

# show results
print(rnase_correlations)

rnase_cor_long <- pivot_longer(rnase_correlations, 
                                cols = starts_with("rnase"), 
                                names_to = "comparison", 
                                values_to = "correlation")

# Plot
ggplot(rnase_cor_long, aes(x = fraction, y = correlation, color = comparison)) +
  geom_point(size = 3) +
  scale_color_manual(values = c(
    "rnase1_vs_rnase2" = "#c9a0dc",  
    "rnase1_vs_rnase3" = "#A9C1A6",  
    "rnase2_vs_rnase3" = "#F5E1A4"   
  )) +
  theme_minimal() +
  labs(title = "Spearman Correlation of RNase Replicates",
       y = "Spearman ρ",
       x = "Fraction",
       color = "Replicate Comparison") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        plot.title = element_text(hjust = 0.5))
```
Output: correlation amongst the rnase values is sporadically lower compared to the ctrl measurments, still within an acceptable range 
-> threshold for acceptable values at 0.35
-> evaluation is therefore reproducable as well 

# 2. data normalisation
## normalisation
### normalisation per fraction
Goal: receive basis for subsequent normalisation
#### determine columnwise mean
```{r}
# function to compute mean columnwise (except Protein_Name)
compute_column_means <- function(df) {
  # select numeric columns (except Protein_Name)
  numeric_data <- df[, sapply(df, is.numeric)]
  
  # calculate mean coulmnwise
  col_means <- colMeans(numeric_data, na.rm = TRUE)
  
  return(col_means)
}

#apply to our data
ctrl_means <- compute_column_means(df_ctrl)
rnase_means <- compute_column_means(df_rnase)


# save means as dataframes
ctrl_means_df <- data.frame(Sample = names(ctrl_means), Mean = as.numeric(ctrl_means))
rnase_means_df <- data.frame(Sample = names(rnase_means), Mean = as.numeric(rnase_means))

#depict results
head(ctrl_means_df)
head(rnase_means_df)

#check whether there are means == 0 (to calculate normalization factor: mean are denominator -> would create NAs if mean were 0)
sum(ctrl_means_df[, 2] == 0)
sum(rnase_means_df[, 2]==0)

```
### normalisation factor
Goal: Calculate reference values per fraction by computing the mean of the two most similar replicates (out of three) to serve as normalization basis.
```{r}
# create function to calculate the mean of the two most similar replicates (out of three replicates)
mean_of_closest_pair <- function(x) {
  combs <- combn(x, 2)                          # all pairwise combinations
  diffs <- abs(combs[1, ] - combs[2, ])         # calculates the absolute distance
  min_idx <- which.min(diffs)                   # indices of minimal distance replicates
  mean(combs[, min_idx])                        # mean of the most similar pair
}

# alter function so it will be applied to three values at a time moving fractionwise
process_blocks <- function(values_vector) {
  n_blocks <- length(values_vector) / 3         # number of fractions
  result <- numeric(n_blocks)
  
  for (i in 1:n_blocks) {
    block <- values_vector[((i - 1) * 3 + 1):(i * 3)] #values vector contains the three values of the replicates for each fraction
    result[i] <- mean_of_closest_pair(block)
  }
  
  return(result)
}

# extract values from the dataframe 
control_values <- ctrl_means_df$Mean    
rnase_values   <- rnase_means_df$Mean

# apply to our data
control_means <- process_blocks(control_values)
rnase_means   <- process_blocks(rnase_values)

# save result as dataframe
closest_means_ctrl <- data.frame(
  Fraktion = paste0("Fraction", 1:length(control_means)),  
  Control_most_similar_pair_mean  = control_means
)

closest_means_rnase <- data.frame(
  Fraktion = paste0("Fraction", 1:length(control_means)),  
  RNase_most_similar_pair_mean = rnase_means)

# depict result
print(closest_means_ctrl)
print(closest_means_rnase)
```
#### create normalisation matrix
Goal: Create normalization factors for each replicate by dividing the reference mean by the replicate value.
```{r}
n_fractions <- 25
n_replicates <- 3

# ctrl_means_df and rnase_means_df contain all replicate means in one column, e.g. 'Mean'
# ctrl_final_means_df and rnase_final_means_df contain one final mean per fraction in 'FinalMean'

# create matrices for replicates of each fraction (25 rows × 3 columns)
ctrl_mat <- matrix(ctrl_means_df$Mean, nrow = n_fractions, ncol = n_replicates, byrow = TRUE)
rnase_mat <- matrix(rnase_means_df$Mean, nrow = n_fractions, ncol = n_replicates, byrow = TRUE)

# initialize normalization matrices with NA values
norm_ctrl_mat <- matrix(NA, nrow = n_fractions, ncol = n_replicates)
norm_rnase_mat <- matrix(NA, nrow = n_fractions, ncol = n_replicates)

# loop over all fractions to calculate normalization factors
for (r in 1:n_fractions) {
  control_values <- ctrl_mat[r, ]  # 3 replicates for Control
  rnase_values <- rnase_mat[r, ]   # 3 replicates for RNase
  
  # retrieve FinalMean from control and RNase data frames
  mean_control <- closest_means_ctrl[r, 2]
  mean_rnase <- closest_means_rnase[r, 2]
  
  # ensure exactly one value per fraction
  if(length(mean_control) != 1 | length(mean_rnase) != 1) {
    stop(paste("More or less than one FinalMean found for fraction", r))
  }
  
  # calculate normalization factors by dividing final mean by replicate values
  norm_ctrl_mat[r, ] <- mean_control / control_values
  norm_rnase_mat[r, ] <- mean_rnase / rnase_values
}

# show first few rows of normalization matrices
head(norm_ctrl_mat)
head(norm_rnase_mat)

# check for any NAs in normalization matrices
sum(is.na(norm_ctrl_mat))
sum(is.na(norm_rnase_mat))
```
#### Apply Normalisation Matrix to Data
Goal: Compensate deviations between replicates and the reference value by multiplying with normalisation factors
```{r}
normalize_df <- function(df, norm_mat) {
  # Extract numeric data excluding the first column ("Protein_Name")
  data_mat <- as.matrix(df[, -1])
  
  # Get the number of blocks (rows) and columns per block from normalization matrix
  n_blocks <- nrow(norm_mat)
  cols_per_block <- ncol(norm_mat)
  
  # Apply normalization blockwise:
  # The normalization matrix is iterated column-wise and applied in this order to the columns of the original data
  for (i in 1:n_blocks) {
    for (j in 1:cols_per_block) {
      col_index <- (i - 1) * cols_per_block + j
      data_mat[, col_index] <- data_mat[, col_index] * norm_mat[i, j]
    }
  }
  
  # Combine Protein_Name column with normalized data and return as a dataframe
  result_df <- cbind(Protein_Name = df$Protein_Name, as.data.frame(data_mat))
  return(result_df)
}

# Apply normalization to control and RNase dataframes
normed_ctrl_df  <- normalize_df(df_ctrl, norm_ctrl_mat)
normed_rnase_df <- normalize_df(df_rnase, norm_rnase_mat)

# Display the first few rows of the normalized dataframes
head(normed_ctrl_df)
head(normed_rnase_df)
```
### Moving Average Filter & Sorting Data
Goal: Apply a moving average filter (window = 3) across fractions to smooth intensity values.
-> Smoothing is applied per replicate and protein; first and last fraction values are kept unchanged.
-> The resulting dataframe is then reordered: columns are grouped by fraction (not by replicate).
```{r}
# Mean filter function for a row (preserving first and last value)
smooth_and_reorder <- function(df, n_fractions = 25, n_replicates = 3) {
  proteins <- df[, 1]
  
  smoothed_reps <- lapply(1:n_replicates, function(rep) {
    cols <- seq(1 + rep, 1 + rep + (n_fractions - 1) * 3, by = 3)
    rep_data <- df[, cols]
    
    smoothed_data <- matrix(NA, nrow = nrow(rep_data), ncol = n_fractions)
    smoothed_data[, 1] <- rep_data[, 1]
    smoothed_data[, n_fractions] <- rep_data[, n_fractions]
    
    for (i in 2:(n_fractions - 1)) {
      smoothed_data[, i] <- rowMeans(rep_data[, (i - 1):(i + 1)], na.rm = TRUE)
    }
    
    smoothed_df <- as.data.frame(smoothed_data)
    colnames(smoothed_df) <- paste0("Fraction", 1:n_fractions, "_Rep", rep)
    smoothed_df
  })
  
  combined_df <- do.call(cbind, smoothed_reps)
  
  # sort columns: first all replicates of first fraction, then fraction 2, etc.
  sorted_cols <- unlist(lapply(1:n_fractions, function(frac) {
    paste0("Fraction", frac, "_Rep", 1:n_replicates)
  }))
  
  final_df <- cbind(Protein = proteins, combined_df[, sorted_cols])
  return(final_df)
}

# apply to our data
final_smoothed_ctrl <- smooth_and_reorder(normed_ctrl_df)
final_smoothed_rnase <- smooth_and_reorder(normed_rnase_df)

# depict result
head(final_smoothed_ctrl)
head(final_smoothed_rnase)

#check if any NAs were created
sum(is.na(final_smoothed_ctrl))
sum(is.na(final_smoothed_rnase))
```
#### checking for replicates with rowsum == 0 
```{r}
# check for any rows with rowSum == 0 (per replicate)
check_zero_replicates_detailed <- function(df) {
  # Remove first column (protein names)
  data_only <- df[, -1]
  protein_names <- df[, 1]
  
  # Initialize lists to store results
  zero_count <- integer(nrow(data_only))
  zero_replicates <- character(nrow(data_only))
  
  # Loop through each row (protein)
  for (i in 1:nrow(data_only)) {
    affected_reps <- c()  # Temporary holder for replicate names
    
    for (rep in 1:3) {
      # Select columns for this replicate across all fractions
      replicate_indices <- seq(from = rep, to = ncol(data_only), by = 3)
      
      # Calculate sum for this replicate
      rep_sum <- sum(data_only[i, replicate_indices])
      
      # If sum is zero, store the replicate name
      if (rep_sum == 0) {
        affected_reps <- c(affected_reps, paste0("Replicate", rep))
      }
    }
    
    # Store number of zero replicates and their names
    zero_count[i] <- length(affected_reps)
    zero_replicates[i] <- if (length(affected_reps) > 0) paste(affected_reps, collapse = ", ") else NA
  }
  
  # Create final summary data frame
  summary_df <- data.frame(
    Protein = protein_names,
    Zero_Replicate_Count = zero_count,
    Affected_Replicates = zero_replicates,
    stringsAsFactors = FALSE
  )
  
  # Optional: filter to only those with at least one zero replicate
  summary_df <- summary_df[summary_df$Zero_Replicate_Count > 0, ]
  
  return(summary_df)
}

# Run for control
ctrl_zero_summary <- check_zero_replicates_detailed(final_smoothed_ctrl)

# Run for RNase
rnase_zero_summary <- check_zero_replicates_detailed(final_smoothed_rnase)

# View examples
head(ctrl_zero_summary)
head(rnase_zero_summary)
```
## Normalisation of Total Amount per Replicate to 100%
Goal: Obtain normalized values per replicate
```{r}
#normalize total amount per replicate to 100%
calc_fraction_percentages <- function(df, zero_info, n_fractions = 25, n_replicates = 3) {
  proteins <- df[, 1]
  percent_reps <- list()
  
  for (rep in 1:n_replicates) {
    # Select replicate columns for all fractions (e.g., Rep1 of all fractions)
    cols <- seq(1 + rep, 1 + rep + (n_fractions - 1) * 3, by = 3)
    rep_data <- as.matrix(df[, cols])
    
    # Get logical vector: TRUE if replicate sum == 0 (should NOT be normalized)
    rep_name <- paste0("Replicate", rep)
    skip_replicate <- zero_info[[rep_name]] == TRUE  # original values are kept

    
    # Calculate row sums
    row_sums <- rowSums(rep_data, na.rm = TRUE)
    
    # Normalize only where rowSum != 0
    percent_data <- rep_data  # start with original values
    
    # Normalize where allowed (i.e., rowSum != 0)
    normalize_rows <- !skip_replicate
    percent_data[normalize_rows, ] <- sweep(
      rep_data[normalize_rows, ],
      1,
      row_sums[normalize_rows],
      FUN = "/"
    ) * 100
    
    # Create dataframe and set column names
    percent_df <- as.data.frame(percent_data)
    colnames(percent_df) <- paste0("Fraction", 1:n_fractions, "_Rep", rep)
    percent_reps[[rep]] <- percent_df
  }
  
  # Combine all replicate dataframes
  combined_percent_df <- do.call(cbind, percent_reps)
  
  # Reorder columns: Fraction1_Rep1, Fraction1_Rep2, Fraction1_Rep3, etc.
  sorted_cols <- unlist(lapply(1:n_fractions, function(frac) {
    paste0("Fraction", frac, "_Rep", 1:n_replicates)
  }))
  
  final_percent_df <- cbind(Protein = proteins, combined_percent_df[, sorted_cols])
  return(final_percent_df)
}


# Run the replicate-zero check
ctrl_zero_info <- check_zero_replicates_detailed(final_smoothed_ctrl)
rnase_zero_info <- check_zero_replicates_detailed(final_smoothed_rnase)

# Run the adjusted percentage calculation
percent_ctrl_df <- calc_fraction_percentages(final_smoothed_ctrl, ctrl_zero_info)
percent_rnase_df <- calc_fraction_percentages(final_smoothed_rnase, rnase_zero_info)

# View results
head(percent_ctrl_df)
head(percent_rnase_df)

#check for NAs
sum(is.na(percent_ctrl_df))
sum(is.na(percent_rnase_df))
```
## Calculate Mean Over the Three Replicates per Fraction
Goal: Obtain normalized values averaged across replicates to get average distribution per fraction
```{r}
calculate_mean_per_fraction <- function(df, n_fractions = 25, n_replicates = 3) {
  proteins <- df[, 1]  # Protein-Names
  data_only <- df[, -1]  # only numeric values

  mean_matrix <- matrix(NA, nrow = nrow(data_only), ncol = n_fractions)

  for (f in 1:n_fractions) {
    # columns for fraction f (3 replicates each)
    cols <- ((f - 1) * n_replicates + 1):(f * n_replicates)
    mean_matrix[, f] <- rowMeans(data_only[, cols], na.rm = TRUE)
  }

  # save result as dataframe
  mean_df <- as.data.frame(mean_matrix)
  colnames(mean_df) <- paste0("Fraction", 1:n_fractions)
  final_df <- cbind(Protein = proteins, mean_df)

  return(final_df)
}

#apply to our data
mean_ctrl_df <- calculate_mean_per_fraction(percent_ctrl_df)
mean_rnase_df <- calculate_mean_per_fraction(percent_rnase_df)

# depict results
head(mean_ctrl_df)
head(mean_rnase_df)

#check if NAs were created
sum(is.na(mean_ctrl_df))
sum(is.na(mean_rnase_df))
```
## Normalisation of Amount of Protein to Equal 100%
Goal: Obtain comparable relative distributions of protein amounts across fractions
```{r}
# normalization of amount of protein to equal 100%

calculate_fraction_percentages <- function(mean_df) {
  proteins <- mean_df[, 1]
  data_only <- mean_df[, -1]
  
  # calculate row sum (sum across all fractions for each protein)
  row_totals <- rowSums(data_only, na.rm = TRUE)
  
  # transform amount into percent
  percent_matrix <- sweep(data_only, 1, row_totals, FUN = "/") * 100
  
  # combine result to a dataframe
  percent_df <- cbind(Protein = proteins, as.data.frame(percent_matrix))
  colnames(percent_df) <- colnames(mean_df)
  
  return(percent_df)
}


# apply to our data
normalized_ctrl_df  <- calculate_fraction_percentages(mean_ctrl_df)
normalized_rnase_df <- calculate_fraction_percentages(mean_rnase_df)

# depict result
head(normalized_ctrl_df)
head(normalized_rnase_df)

#check if any NAs were created
sum(is.na(normalized_ctrl_df))
sum(is.na(normalized_rnase_df))
```
## Check for Empty Protein Distributions After Normalization
Goal: Identify proteins with zero total abundance after normalization
```{r}
# For Control data
# Calculate row sums excluding the first column ("Protein_Name")
row_sums_ctrl <- rowSums(normalized_ctrl_df[, -1])

# Create a boolean vector: TRUE where row sum equals zero
rows_with_zero_sum_ctrl <- row_sums_ctrl == 0

# Check if any rows have a sum of zero
any(rows_with_zero_sum_ctrl)

# For RNase data
# Calculate row sums excluding the first column ("Protein_Name")
row_sums_rnase <- rowSums(normalized_rnase_df[, -1])

# Create a boolean vector: TRUE where row sum equals zero
rows_with_zero_sum_rnase <- row_sums_rnase == 0

# Check if any rows have a sum of zero
any(rows_with_zero_sum_rnase)
```
# Visualization of the distribution of any protein before and after rnase treatment
```{r}
# choose a protein and extract data across the fractions
protein_id <- "KIF2A_HUMAN"
protein_row <- which(normalized_ctrl_df[, 1] == protein_id)
intensities_ctrl <- as.numeric(normalized_ctrl_df[protein_row, 2:26])
intensities_rnase <- as.numeric(normalized_rnase_df[protein_row, 2:26])
fraktionen <- 1:25

# prepare the plot
plot(fraktionen, intensities_ctrl, type = "n",
     xlab = "Fractions (1–25)", ylab = "Amount Of Protein",
     main = paste("Comparison: Control and RNase for", protein_id),
     ylim = c(0, max(c(intensities_ctrl, intensities_rnase)) * 1.1),
     xaxt = "n")
axis(1, at = 1:25)
grid(col = "lightgray", lty = "dotted")

hex_to_rgba <- function(hex_code, alpha = 1) {
  hex_code_clean <- substring(hex_code, 2)
  r <- strtoi(substr(hex_code_clean, 1, 2), 16L) / 255
  g <- strtoi(substr(hex_code_clean, 3, 4), 16L) / 255
  b <- strtoi(substr(hex_code_clean, 5, 6), 16L) / 255
  rgb(r, g, b, alpha)
}

ctrl_fill_color <- hex_to_rgba("#A9C1A6", 0.3)
ctrl_line_color <- hex_to_rgba("#A9C1A6", 1)
rnase_fill_color <- hex_to_rgba("#c9a0dc", 0.3)
rnase_line_color <- hex_to_rgba("#c9a0dc", 1)


# control graph
polygon(c(fraktionen, rev(fraktionen)),
        c(rep(0, length(fraktionen)), rev(intensities_ctrl)),
        col = ctrl_fill_color, border = NA)
lines(fraktionen, intensities_ctrl, col = ctrl_line_color, lwd = 2)

# rnase graph 
polygon(c(fraktionen, rev(fraktionen)),
        c(rep(0, length(fraktionen)), rev(intensities_rnase)),
        col = rnase_fill_color, border = NA)
lines(fraktionen, intensities_rnase, col = rnase_line_color, lwd = 2)

legend("topright",
       legend = c("Control", "RNase"),
       col = c(ctrl_line_color, rnase_line_color),
       lwd = 2,
       bg = "white")


# visualize arrows for better depiction of the shift
# find the global maxima
max_ctrl_idx <- which.max(intensities_ctrl)
max_ctrl_val <- intensities_ctrl[max_ctrl_idx]
max_rnase_idx <- which.max(intensities_rnase)
max_rnase_val <- intensities_rnase[max_rnase_idx]

# add arrows for the amplitude of the maxima
arrows(x0 = fraktionen[max_ctrl_idx], y0 = 0,
       x1 = fraktionen[max_ctrl_idx], y1 = max_ctrl_val,
       col = ctrl_line_color, lwd = 1.5, length = 0.1)

arrows(x0 = fraktionen[max_rnase_idx], y0 = 0,
       x1 = fraktionen[max_rnase_idx], y1 = max_rnase_val,
       col = rnase_line_color, lwd = 1.5, length = 0.1)

text(fraktionen[max_ctrl_idx], max_ctrl_val + 0.08 * max_ctrl_val,
     labels = "Amplitude", col = ctrl_line_color, cex = 0.8)

text(fraktionen[max_rnase_idx], max_rnase_val + 0.08 * max_rnase_val,
     labels = "Amplitude", col = rnase_line_color, cex = 0.8)

# improved position of the arrow 
arrow_y <- min(max_ctrl_val, max_rnase_val)

# shift distance arrow
arrows(x0 = fraktionen[max_ctrl_idx], y0 = arrow_y,
       x1 = fraktionen[max_rnase_idx], y1 = arrow_y,
       col = "black", lwd = 1.6, angle = 15, length = 0.1, code = 2)
text(x = mean(c(fraktionen[max_ctrl_idx], fraktionen[max_rnase_idx])),
     y = arrow_y + 0.08 * arrow_y,
     labels = "Shift Distance", col = "black", cex = 0.8)

```

# visualize normalisation
```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(scales)

# define color
ctrl_fill_color <- alpha("#A9C1A6", 0.3)
ctrl_line_color <- alpha("#A9C1A6", 1)
rnase_fill_color <- alpha("#c9a0dc", 0.3)
rnase_line_color <- alpha("#c9a0dc", 1)

# 1. Convert raw data to long format
ctrl_long <- df_ctrl %>%
  pivot_longer(cols = -Protein_Name, names_to = "Sample", values_to = "Intensity") %>%
  separate(Sample, into = c("Fraction", "Condition", "Replicate"), sep = "_", extra = "drop") %>%
  mutate(
    Fraction = as.integer(sub("Fraction", "", Fraction)),
    Condition = "Ctrl",
    Status = "Raw",
    Intensity = as.numeric(Intensity)
  )

rnase_long <- df_rnase %>%
  pivot_longer(cols = -Protein_Name, names_to = "Sample", values_to = "Intensity") %>%
  separate(Sample, into = c("Fraction", "Condition", "Replicate"), sep = "_", extra = "drop") %>%
  mutate(
    Fraction = as.integer(sub("Fraction", "", Fraction)),
    Condition = "RNase",
    Status = "Raw",
    Intensity = as.numeric(Intensity)
  )

raw_all <- bind_rows(ctrl_long, rnase_long)

# 2. Convert normalized data to long format
norm_ctrl_long <- normalized_ctrl_df %>%
  pivot_longer(cols = -Protein, names_to = "Fraction", values_to = "Intensity") %>%
  mutate(
    Fraction = as.integer(sub("Fraction", "", Fraction)),
    Condition = "Ctrl",
    Status = "Normalized",
    Intensity = as.numeric(Intensity)
  ) %>%
  rename(Protein_Name = Protein)

norm_rnase_long <- normalized_rnase_df %>%
  pivot_longer(cols = -Protein, names_to = "Fraction", values_to = "Intensity") %>%
  mutate(
    Fraction = as.integer(sub("Fraction", "", Fraction)),
    Condition = "RNase",
    Status = "Normalized",
    Intensity = as.numeric(Intensity)
  ) %>%
  rename(Protein_Name = Protein)

norm_all <- bind_rows(norm_ctrl_long, norm_rnase_long)

# 3. Combine both datasets
combined <- bind_rows(raw_all, norm_all)

# 4. Determine smallest positive value (for offset)
min_positive <- combined %>%
  filter(!is.na(Intensity), Intensity > 0) %>%
  summarise(min_value = min(Intensity)) %>%
  pull(min_value)

# 5. Replace zero and NA values with the smallest positive value
# Then calculate shifted log10 values (smallest value set to 0)
combined_shifted <- combined %>%
  mutate(
    Intensity_fixed = ifelse(is.na(Intensity) | Intensity <= 0, min_positive, Intensity),
    Intensity_shifted = log10(Intensity_fixed) - log10(min_positive)
  )

# 6. Filter only the first 10 proteins
combined_10_shifted <- combined_shifted %>%
  filter(Protein_Name %in% unique(Protein_Name)[1:10])

# 7. Plot with shifted log10 scale
ggplot(combined_10_shifted, aes(x = Fraction, y = Intensity_shifted, color = Status)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c(
    "Raw" = ctrl_line_color,
    "Normalized" = rnase_line_color
  )) +
  facet_grid(Condition ~ Protein_Name) +
  labs(title = "Proportionally shifted log10 intensities (min value set to 0)",
       x = "Fraction", y = "Shifted log10 Intensity") +
  theme_minimal(base_size = 8) +
  theme(
    axis.text.x = element_text(size = 5, angle = 45, hjust = 1),     # Fraction numbers smaller & angled
    axis.text.y = element_text(size = 6),                            # Y-axis text smaller
    axis.title = element_text(size = 7),                            # Axis titles smaller
    strip.text.x = element_text(size = 5, margin = margin(b = 3)),   # Protein names (x facets) smaller
    strip.text.y = element_text(size = 6, margin = margin(r = 3)),   # y facets smaller
    legend.text = element_text(size = 6),                           # Legend text smaller
    legend.title = element_text(size = 7),                          # Legend title smaller
    panel.spacing.x = unit(0.5, "lines"),                           # Less spacing between x facets
    panel.spacing.y = unit(0.5, "lines")                            # Less spacing between y facets
  )
```


###Batch Effect###
Goal: Determine if there is a batch effect and remove it. Batch effect in this case refers to differences in data not related to biological reasons but rather technological faults. 

##CONTROL


##load data and creat matrix
```{r}
# 1. Extract expression data and save it as matrix, delete protein names
ctrl_expr_matrix <- as.matrix(percent_ctrl_df[, -1])  
```

##create batch vector 
```{r}
# 1. Extract sample names to create batch labels 
LC_sample_names <- colnames(ctrl_expr_matrix)

# 2. Create batch labels out of sample names 
#create batches from replicates in order to have comparable batches
LC_batch <- sub(".*_(Rep\\d)", "\\1", LC_sample_names)
LC_batch <- factor(LC_batch, levels = c("Rep1", "Rep2", "Rep3"))

# 3. Assign colors 
LC_batch_colors <- c("Rep1" = "#c9a0dc", "Rep2" = "#A9C1A6", "Rep3" = "#F5E1A4")
colors <- LC_batch_colors[as.character(LC_batch)]

#4.shorten labels 
short_labels <- gsub("Fraction([0-9]+)_Rep([0-9])", "F\\1_R\\2", colnames(ctrl_expr_matrix))

```

##boxplot ctrl 
Goal:
Visualize data in a boxplot in order to observe first differences 
```{r}

#1. create boxplot
boxplot(ctrl_expr_matrix,
        main = "Expression by sample (batch coloured), ctrl",
        ylab = "Expression",
        xlab = "Samples",
        las = 2,                     
        col = LC_batch_colors,
        outline = FALSE,
        cex.axis = 0.6,
        names = short_labels)            

#2. add key
legend_labels <- levels(LC_batch)
legend_colors <- LC_batch_colors[legend_labels]

legend("topright",
       legend = legend_labels,
       fill = legend_colors,
       title = "Batch")

```


# interpretation 
there is no distinct differences between replicates, which rules out a servere batch effect  

```{r}
head(ctrl_expr_matrix)
```

```{r}##
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("limma")


```##

##limma ctrl
first trial of removing the batch effect using limma

```{r}


library(limma)


#1. apply a log 2 transformation 
# necessary for limma, stabilizes variance, improves interpretability , reduces outlier effects 

#add +1 in order to prevent log2(0)

ctrl_expr_matrix_log <- log2(ctrl_expr_matrix + 1)

#2. Correction with limma
L_ctrl_expr_corrected <- removeBatchEffect(ctrl_expr_matrix_log, batch = batch)



#3. draw boxplot beofre correction 
#use log2 transformed matrix for comparison 
# expectation
    #before: samples of one batch have similar median, but differences in batches
 #boxplot before
boxplot(ctrl_expr_matrix_log,
        main = "Expression before limma correction (batch coloured), ctrl",
        ylab = "Expression",
        xlab = "Samples",
        las = 2,                     
        col = LC_batch_colors,
        outline = FALSE,
        cex.axis = 0.6,
        names = short_labels) 
#add key
legend_labels <- c("Rep1", "Rep2", "Rep3")
legend_colors <- LC_batch_colors[legend_labels]
legend("topright",
       legend = legend_labels,
       fill = legend_colors,
       title = "Batch")

#4.Boxplot after correction
boxplot(L_ctrl_expr_corrected,
        main = "Expression after limma correction (batch coloured), ctrl",
        ylab = "Expression",
        xlab = "Samples",
        las = 2,                    
        col = LC_batch_colors,
        outline = FALSE,
        cex.axis = 0.6,
        names = short_labels)             


# add key
legend_labels <- c("Rep1", "Rep2", "Rep3")
legend_colors <- LC_batch_colors[legend_labels]
legend("topright",
       legend = legend_labels,
       fill = legend_colors,
       title = "Batch")
```

#Interpretation 
There is no severe change visible 


## Batch effect pca limma ctrl
#expectation
before: clusters of the same batch, strong separation along principle components 
after: no separation visible
```{r}
#1. Prepare PCA
LC_pca_before <- prcomp(t(ctrl_expr_matrix_log), scale. = TRUE)
LC_pca_after  <- prcomp(t(L_ctrl_expr_corrected), scale. = TRUE)
# Transposing the matrix is necessary because the observations and variables are swapped

#2. plot

# next to each other
par(mfrow = c(1, 2))



# PCA before correction
plot(LC_pca_before$x[, 1:2],
     col = LC_batch_colors,
     pch = 19,
     main = "PCA before limma correction, ctrl")
legend("topright", legend = levels(LC_batch), col = LC_batch_colors, pch = 19)

# PCA after correction
plot(LC_pca_after$x[, 1:2],
     col = LC_batch_colors,
     pch = 19,
     main = "PCA after limma correction, ctrl")
legend("topright", legend = levels(LC_batch), col = LC_batch_colors, pch = 19)


```



#interpretation
There is no severe change visible, also there is no distinct differences between replicates

##heatmap
#expectation: 
before: samples are grouped by batch, there is a clear separation of batches shown in the annotation 
after: samples are grouped according to biological features, no separation 


```{r}
#prep: Set short labels as new column names
colnames(L_ctrl_expr_corrected) <- short_labels
colnames(ctrl_expr_matrix_log) <- short_labels

#1. create annotation for Heatmap
annotation_col <- data.frame(LC_Batch = LC_batch)
rownames(annotation_col) <- colnames(ctrl_expr_matrix_log)

#2. 
library(pheatmap)

# heatmap before correction
pheatmap(ctrl_expr_matrix_log,
         annotation_col = annotation_col,
         annotation_colors = list(LC_Batch = LC_batch_colors),
         main = "Heatmap before limma batch correction, ctrl",
         fontsize_col=5)



# heatmap after correction
pheatmap(L_ctrl_expr_corrected,
         annotation_col = annotation_col,
         annotation_colors = list(LC_Batch = LC_batch_colors),
         main = "Heatmap after limma batch correction, ctrl ",
         fontsize_col=5)
```



#interpretation 
before the correction there is no major separation visible, after the correction there is a slight change visible 


##protein names back
```{r}

#add column with protein names again 
L_ctrl_expr_corrected_df <- data.frame(
  Protein = percent_ctrl_df[, 1],  
L_ctrl_expr_corrected,               
  check.names = FALSE                  # Prevents renaming of columns
)

colnames(L_ctrl_expr_corrected_df)
```

```{r}
head(L_ctrl_expr_corrected_df)
```

##RNASE

##boxplot rnase
Goal:
Visualize data in a boxplot in order to observe first differences 
##load data and creat matrix
```{r}
# 1. Extract expression data and save it as matrix, delete protein names
rnase_expr_matrix <- as.matrix(percent_ctrl_df[, -1])  
```

##create batch vector rnase limma
```{r}

#1. Extract sample names to create batch labels 
LR_sample_names <- colnames(rnase_expr_matrix)

#2. Create batch labels out of sample names 
#create batches from replicates in order to have comparable batches
LR_batch <- sub(".*_(Rep\\d)", "\\1", LR_sample_names)
LR_batch <- factor(LR_batch, levels = c("Rep1", "Rep2", "Rep3"))

#3. Assign colors 
LR_batch_colors <- c("Rep1" = "#c9a0dc", "Rep2" = "#A9C1A6", "Rep3" = "#F5E1A4")
colors <- LR_batch_colors[as.character(LR_batch)]

#4.shorten labels 
short_labels <- gsub("Fraction([0-9]+)_Rep([0-9])", "F\\1_R\\2", colnames(rnase_expr_matrix))

```


##boxplot 
Goal:
Visualize data in a boxplot in order to observe first differences 
```{r}

#1. draw boxplot
boxplot(rnase_expr_matrix,
        main = "Expression by sample (batch coloured), rnase",
        ylab = "Expression",
        xlab = "Samples",
        las = 2,                     
        col = LR_batch_colors,
        outline = FALSE,
        cex.axis = 0.6,
        names = short_labels)              

#2. add key
legend_labels <- c("Rep1", "Rep2", "Rep3")
legend_colors <- LR_batch_colors[legend_labels]
legend("topright",
       legend = names(LR_batch_colors),
       fill = legend_colors,
       title = "Batch")
```

#interpretation 
there is no distinct differences between replicates, which rules out a servere batch effect 


```{r}
head(rnase_expr_matrix)
```

#limma rnase
first trial of removing the batch effect using limma
```{r}
library(limma)

#1. Apply a log 2 transformation 
# necessary for limma, stabilizes variance, improves interpretability, reduces outlier effects 
#add +1 in order to prevent log2(0)
rnase_expr_matrix_log <- log2(rnase_expr_matrix + 1)

#2. Correction with limma
L_rnase_expr_corrected <- removeBatchEffect(rnase_expr_matrix_log, LR_batch = LR_batch)

#3. Plot
#use log2-transformed matrix for comparison
# expectation
    #before: samples of one batch have similar median, but differences in batches
# Boxplot before correction

boxplot(rnase_expr_matrix_log,
        main = "Expression before limma correction (batch coloured, RNAse)",
        ylab = "Expression",
        xlab = "Samples",
        las = 2,                      
        col = colors,
        outline = FALSE,
        cex.axis = 0.6,
        names=short_labels)             

#add key 
legend_labels <- c("Rep1", "Rep2", "Rep 3")
legend("topright",
       legend = legend_labels,
       fill = LR_batch_colors,
       title = "Batch",
       cex=0.4)


# Boxplot after correction
boxplot(L_rnase_expr_corrected,
        main = "Expression after limma correction (batch coloured, RNAse)",
        ylab = "Expression",
        xlab = "Samples",
        las = 2,                      
        col = colors,
        outline = FALSE,
        cex.axis = 0.6,
        names=short_labels)             


#add key
llegend_labels <- c("Rep1", "Rep2", "Rep 3")
legend("topright",
       legend = legend_labels,
       fill = LR_batch_colors,
       title = "Batch",
       cex=0.4)
```

#Interpretation 
There is no severe change visible  

#more visualizing

## Batch effect pca limma rnase
#expectation
before: clusters of the same batch, strong separation along principle components 
after: no separation visible
```{r}
#1. Prepare PCA
LR_pca_before <- prcomp(t(rnase_expr_matrix_log), scale. = TRUE)
LR_pca_after  <- prcomp(t(L_rnase_expr_corrected), scale. = TRUE)
# Transposing the matrix is necessary because the observations and variables are swapped

#2. plot
#plot next to eachother
par(mfrow = c(1, 2))


# PCA before correction
plot(LR_pca_before$x[, 1:2],
     col = colors,
     pch = 19,
     main =  "PCA before limma correction, RNAse", 
     cex.main=0.8)
legend("topright", legend = levels(LR_batch), col = colors, pch = 19)

# PCA after correction
plot(LR_pca_after$x[, 1:2],
     col = colors,
     pch = 19,
     main = "PCA after limma correction, RNAse", 
     cex.main=0.8)
legend("topright", legend = levels(LR_batch), col = colors, pch = 19)

```

#interpretation
There is no severe change visible, also there are no distinct differences between replicates, which rules out a severe batch effect 



##heatmap rnase limma
#expectation: 
Before: Samples are grouped by batch, and there is a clear separation of batches shown in the annotation 
after: samples are grouped according to biological feature, no separation

```{r}
#prep: Set short labels as new column names
colnames(L_rnase_expr_corrected) <- short_labels
colnames(rnase_expr_matrix_log) <- short_labels

#1. create annotation for Heatmap
annotation_col <- data.frame(LC_Batch = LC_batch)
rownames(annotation_col) <- colnames(ctrl_expr_matrix_log)

#2. 
library(pheatmap)

# heatmap before correction
pheatmap(rnase_expr_matrix_log,
         annotation_col = annotation_col,
         annotation_colors = list(LC_Batch = LC_batch_colors),
         main = "Heatmap before limma batch correction, RNAse",
         fontsize_col=5)



# heatmap after correction
pheatmap(L_rnase_expr_corrected,
         annotation_col = annotation_col,
         annotation_colors = list(LC_Batch = LC_batch_colors),
         main = "Heatmap after limma batch correction, RNAse",
         fontsize_col=5)

```



#interpretation 
no change visible 



```{r}
#add column with protein names again 
rnase_expr_corrected_df <- data.frame(
  Protein = percent_ctrl_df[, 1],  
  L_rnase_expr_corrected,                
  check.names = FALSE                 # Prevents renaming of columns
)
```


```{r}
head(rnase_expr_corrected_df)
```



##SVA
SVA is used as alternative packages. SVA is a more flexible package as it does not need any exact batch annotation and can detect more hidden disruptive factors. For experiments with many replicates and fractions this package appears more suitable 


```{r}##
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("sva")
```


```{r}##
install.packages("sva")
```

##control SVA
```{r}

# load libraries 
library(sva)
library(stringr)
library(limma)

#1. Create the expression matrix without column of protein names 
ctrl_expr_matrix <- as.matrix(percent_ctrl_df[, -1])  


#2. create sampleInfo 
ctrl_samples <- colnames(ctrl_expr_matrix)

# Extract "fraction" and "replicate" information out of colnames
fraction <- str_extract(ctrl_samples, "Fraction\\d+")
replicate <- str_extract(ctrl_samples, "Rep\\d+")

# create sampleInfo dataframe
C_sampleInfo <- data.frame(
  fraction = factor(fraction),
  replicate = factor(replicate)
)

# 3.Create a design matrix for the model
#mod full model with known factors
mod <- model.matrix(~ fraction + replicate, data = C_sampleInfo)
#mod0 null model without descriptive factors but standard
mod0 <- model.matrix(~ 1, data = C_sampleInfo)  # Nullmodell ohne Prädiktoren
#models needed in order to estimate surrogate variables which model unexplained variability 

# 4.Execute SVA 
sva_res_ctrl <- sva(ctrl_expr_matrix, mod, mod0)
#SVA checks for hidden variables that can not be explained by replicates or fractions
#returns a list with surrogate variables (SV) that function as batch variables 

# 5. extent design matrix by SVs
mod_sv <- cbind(mod, sva_res_ctrl$sv)

# (6) OPTIONAL:If batch is known, ComBat can be used 
# if batch is known, not the case 
# batch <- sampleInfo$batch_variable
# expression_corrected <- ComBat(dat = expression_matrix, batch = batch, mod = mod)

# 7) Correct data with SV using linear regression 
# lmFit and eBayes are tools from limma used to fit linear regression to expressions data
fit <- lmFit(ctrl_expr_matrix, mod_sv)
fit <- eBayes(fit)

# extract residues: expressions data - modeled effects (fraction + replicate + SVs) 
S_corrected_expr_ctrl <- residuals(fit, ctrl_expr_matrix)

# 'S_corrected_expr_ctrl' contains the corrected dataset 

```



##create batch vector for following visualization
```{r}
# 1. extract sample names to create batch labels 
SC_sample_names <- colnames(ctrl_expr_matrix)
# 2. create batch labels out of sample names 
#create batches from replicates in order to have comparable batches
SC_batch <- sub(".*_(Rep\\d)", "\\1", sample_names)
SC_batch <- factor(SC_batch, levels = c("Rep1", "Rep2", "Rep3"))

# 3. assign colors 
SC_batch_colors <- c("Rep1" = "#c9a0dc", "Rep2" = "#A9C1A6", "Rep3" = "#F5E1A4")
colors <- SC_batch_colors[as.character(SC_batch)]

#4.shorten labels 
SC_short_labels <- gsub("Fraction([0-9]+)_Rep([0-9])", "F\\1_R\\2", colnames(ctrl_expr_matrix))

```

##after and before boxplot
#expectation
before: different levels of medians, span and spread; differences between samples caused by batches
after: reduced differences

```{r}

#1. Boxplot before correction
boxplot(ctrl_expr_matrix,
        main = "Expression before SVA correction (replicate colored, ctrl)",
        ylab = "Expression",
        xlab = "Samples",
        las = 2,
        col = colors,
        outline = FALSE,
        cex.axis = 0.6,
        names=SC_short_labels)
#add key 
legend("topright",
       legend = names(SC_batch_colors),
       fill = SC_batch_colors,
       title = "Replicate",
       cex=0.4)


#2. Boxplot after correction
boxplot(S_corrected_expr_ctrl,
        main = "Expression after SVA correction (replicate colored, ctrl)",
        ylab = "Expression",
        xlab = "Samples",
        las = 2,
        col = colors,
        outline = FALSE,
        cex.axis = 0.6,
        names=SC_short_labels)

# add key 
legend("topright",
       legend = names(SC_batch_colors),
       fill = SC_batch_colors,
       title = "Replicate",
       cex=0.4)


```

#Interpretation 
before: there is some differences between replicates visible but median appears to be on a similar level 
after: big shifts, differences in replicates not as severe 


## visualisierung mittels PCA
#expectations 
before: possible distortion caused by scale differences, principle components show batch separation 
after: no separation visible

```{r}

# 1) PCA vorbereiten
SC_pca_before <- prcomp(t(ctrl_expr_matrix), scale. = TRUE)
SC_pca_after  <- prcomp(t(S_corrected_expr_ctrl), scale. = TRUE)
# Transposing the matrix is necessary because the observations and variables are swapped

#2. Plot: 

#next to each other
par(mfrow = c(1, 2))

#PCA before correction
plot(SC_pca_before$x[, 1:2],
     col = colors,
     pch = 19,
     main = "PCA before SVA correction, ctrl ",
     cex.main=0.8,
     xlab = "PC1", ylab = "PC2")

legend("topright", legend = levels(SC_batch), col = colors, pch = 19, cex=0.4)


# PCA after correction
plot(SC_pca_after$x[, 1:2],
     col = colors,
     pch = 19,
     main = "PCA after SVA correction, ctrl ",
     cex.main=0.8,
     xlab = "PC1", ylab = "PC2")

legend("topright", legend = levels(SC_batch), col = colors, pch = 19, cex=0.4)



```

#Interpretation 
before: No distinct differences between replicates, or separation by PCs
after: Spread is not as severe 


##Heatmap
#expectations
Before: Samples grouped by batch, and there is a clear separation of batches shown in the annotation; outliers, high variability
after: samples are grouped according to biological feature, no separation 

```{r}

#prep: Set short labels as new column names
colnames(S_corrected_expr_ctrl) <- short_labels
colnames(ctrl_expr_matrix) <- short_labels

#1. create annotation for Heatmap
annotation_col <- data.frame(SC_Batch = SC_batch)
rownames(annotation_col) <- colnames(ctrl_expr_matrix)

#2. 
library(pheatmap)

# heatmap before correction
pheatmap(ctrl_expr_matrix,
         annotation_col = annotation_col,
         annotation_colors = list(SC_Batch = SC_batch_colors),
         main = "Heatmap before SVA batch correction, ctrl",
         fontsize_col=5)



# heatmap after correction
pheatmap(S_corrected_expr_ctrl,
         annotation_col = annotation_col,
         annotation_colors = list(SC_Batch = SC_batch_colors),
         main = "Heatmap after batch SVA correction, ctrl",
         fontsize_col=5)


```

#Interpretation
before: there is no extreme clustering but still some visible 
after: slight scale difference, difference in clustering 



##create df with protein column
```{r}
#add column with protein names again 
S_corrected_expr_ctrl_df <- data.frame(
  Protein = percent_ctrl_df[, 1],  
  S_corrected_expr_ctrl,                
  check.names = FALSE                 # Prevents renaming of columns
)
```



```{r}
head(S_corrected_expr_ctrl_df)
```
 




## SVA RNASE

```{r}


# load libraries 
library(sva)
library(stringr)
library(limma)

#1. create expression matrix without column of protein names 
rnase_expr_matrix <- as.matrix(percent_rnase_df[, -1])

# 2.create sampleInfo 
rnase_samples <- colnames(rnase_expr_matrix)

# Extract "fraction" and "replicate" information out of colnames
rnase_fraction <- str_extract(rnase_samples, "Fraction\\d+")
rnase_replicate <- str_extract(rnase_samples, "Rep\\d+")

# create sampleInfo dataframe
sampleInfo_R <- data.frame(
  rnase_fraction = factor(rnase_fraction),
  rnase_fraction = factor(rnase_fraction)
)

# 3. Create a design matrix for the model
#mod full model with known factors
mod <- model.matrix(~ fraction + replicate, data = sampleInfo_R)
#mod0 null model without descriptive factors but standard
mod0 <- model.matrix(~ 1, data = sampleInfo_R)  
#models needed in order to estimate surrogate variables that model unexplained variability 

# 4. Execute SVA 
sva_res_rnase <- sva(rnase_expr_matrix, mod, mod0)
#SVA checks for hidden variables that can be explained by replicates or fractions
#returns a list with surrogate variables (SV) that function as batch variables 


# 5. extent design matrix with SVs
R_mod_sv <- cbind(mod, sva_res_rnase$sv)

# (6) OPTIONAL:If batch is known, ComBat can be used 
# if batch is known, not the case 
# batch <- sampleInfo$batch_variable
# expression_corrected <- ComBat(dat = expression_matrix, batch = batch, mod = mod)

# 7. Correct data with SV using linear regression 
# lmFit and eBayes are tools from limma used to fit linear regression to expression data
fit <- lmFit(rnase_expr_matrix, mod_sv)
fit <- eBayes(fit)

# extract residues: expressions data - modelled effects (fraction + replicate + SVs) 

S_corrected_expr_rnase <- residuals(fit, rnase_expr_matrix)

# S_corrected_expr_rnase contains the corrected dataset

```



```{r}
##create batch vector for following visualization
# 1. extract sample names to create batch labels 
SR_sample_names <- colnames(rnase_expr_matrix)

# 2. create batch labels out of sample names 
#create batches from replicates in order to have comparable batches
SR_batch <- sub(".*_(Rep\\d)", "\\1", sample_names)
SR_batch <- factor(SR_batch, levels = c("Rep1", "Rep2", "Rep3"))

# 2. assign colors
SR_batch_colors <- c("Rep1" = "#c9a0dc", "Rep2" = "#A9C1A6", "Rep3" = "#F5E1A4")
colors <- SR_batch_colors[as.character(SR_batch)]

#4.shorten labels 
short_labels <- gsub("Fraction([0-9]+)_Rep([0-9])", "F\\1_R\\2", colnames(rnase_expr_matrix))

```



## after and before boxplot
#expectation 
before: different levels of medians, span and spread; differences between samples caused by batches
after: reduced differences

```{r}
# 1. Boxplot before correction
boxplot(rnase_expr_matrix,
        main = "Expression before SVA correction (replicate colored, RNAse)",
        ylab = "Expression",
        xlab = "Samples",
        las = 2,
        col = colors,
        outline = FALSE,
        cex.axis = 0.6,
        names=short_labels)
#add key 
legend("topright",
       legend = names(SR_batch_colors),
       fill = SR_batch_colors,
       title = "Replicate",
       cex=0.4)


# 2. Boxplot after correction
boxplot(S_corrected_expr_rnase,
        main = "Expression after SVA correction (replicate colored, RNAse)",
        ylab = "Expression",
        xlab = "Samples",
        las = 2,
        col = colors,
        outline = FALSE,
        cex.axis = 0.6,
        names=short_labels)

# add key 
legend("topright",
       legend = names(SR_batch_colors),
       fill = SR_batch_colors,
       title = "Replicate",
       cex=0.4)


```

#Interpretation 
before: there is some differences between replicates visible but median appears to be on a similar level 
after: big shifts, differences in replicates not as severe 


## visualisierung mittels PCA
#expectations 
Before: clusters of the same batch with distinct separation of clusters; distortion caused by scale differences; principle components show batch separation 
after: no separation visible

```{r}

# 1) prepare PCA 
SR_pca_before <- prcomp(t(rnase_expr_matrix), scale. = TRUE)
SR_pca_after  <- prcomp(t(S_corrected_expr_rnase), scale. = TRUE)
# Transposing the matrix is necessary because the observations and variables are swapped

#2. plot
#next to each other
par(mfrow = c(1, 2))  

#PCA before correction
plot(SR_pca_before$x[, 1:2],
     col = colors,
     pch = 19,
     main = "PCA before SVA correction, RNAse",
     cex.main=0.8,
     xlab = "PC1", ylab = "PC2")

legend("topright", legend = levels(SR_batch), col = batch_colors, pch = 19,cex=0.4)

# PCA after correction
plot(SR_pca_after$x[, 1:2],
     col = colors,
     pch = 19,
     main = "PCA after SVA correction, RNAse ",
     cex.main=0.8,
     xlab = "PC1", ylab = "PC2")
legend("topright", legend = levels(SC_batch), col = colors, pch = 19,cex=0.4)

# 5) reset layout to single plot default
par(mfrow = c(1, 1))

```
#Interpretation 
before: No distinct differences between replicates, no obvious separation by PCs 
after: reduced separation and clustering visi

##Heatmap
#expectations
Before: Samples  grouped by batch, and there is a clear separation of batches shown in the annotation; outliers, high variability  
After: samples are grouped according to biological feature, no separation 

```{r}

#prep: Set short labels as new column names
colnames(S_corrected_expr_rnase) <- short_labels
colnames(rnase_expr_matrix) <- short_labels

# 1. create annotation for Heatmap
annotation_col <- data.frame(SR_Batch = SR_batch)
rownames(annotation_col) <- colnames(rnase_expr_matrix)

# Werte clipping basierend auf Perzentilen
expr_clipped <- S_corrected_expr_rnase
low <- quantile(expr_clipped, 0.01)
high <- quantile(expr_clipped, 0.99)

expr_clipped[expr_clipped < low] <- low
expr_clipped[expr_clipped > high] <- high

#2. 
library(pheatmap)

# heatmap before correction
pheatmap(rnase_expr_matrix,
         annotation_col = annotation_col,
         annotation_colors = list(SR_Batch = SR_batch_colors),
         main = "Heatmap before SVA batch correction, RNAse ",
         fontsize_col=5)



# heatmap after correction
pheatmap(S_corrected_expr_rnase,
         annotation_col = annotation_col,
         annotation_colors = list(SR_Batch = SR_batch_colors),
         main = "Heatmap after SVA batch correction, RNAse ",
         fontsize_col=5)

```


#Interpretation
Before: Clustering is visible in the annotation 
After: Difference in scale, clusters still appear



##create df with protein column
```{r}
#add column with protein names again 
S_corrected_expr_rnase_df <- data.frame(
  Protein = percent_rnase_df[, 1],  
  S_corrected_expr_rnase,                
  check.names = FALSE                 # Prevents renaming of columns
)
```



```{r}
head(S_corrected_expr_rnase_df)
```
 
 
# 3. shift analysis
Goal: Classify proteins as RNA-dependent (RDP) or non-RNA-dependent (non-RDP) based on differences between control and RNase condition
## Tests
Goal: characterize the differences between Ctrl and RNase for each protein
### Visualization of one protein
Goal: get an idea of criteria that can be compared to determine significance of shift
```{r}
#visualize distribution for one protein 
protein_id <- "SPB6_HUMAN"

# extract amount of protein
intensities <- as.numeric(normalized_ctrl_df[4, 2:26 ])
fraktionen <- 1:25

# Plot 
plot(fraktionen, intensities, type = "n",
       xlab = "Fraktion (1–25)", ylab = "Normierte Intensität (Fläche = 1)",
       main = paste("Dichte-Plot von", protein_id),
       ylim = c(0, max(intensities)*1.1))

 polygon(c(fraktionen, rev(fraktionen)),
          c(rep(0, length(fraktionen)), rev(intensities)),
          col = "lavender", border = NA)
  lines(fraktionen, intensities, col = "lavender", lwd = 2)
```

### maxima
Goal: find all local maxima 
```{r}
# IdentifY position of global maxima
# recognize peaks with diff()
find_peaks_diff <- function(x, threshold = 2) {
  n <- length(x)
  peaks <- logical(n)
  
  dx <- diff(x)
  
  # Ensure maximum instead of minimum
  for (i in 2:(n-1)) {
    if (x[i] >= threshold && dx[i-1] > 0 && dx[i] < 0) {
      peaks[i] <- TRUE
    }
  }
 
  if (x[1] >= threshold && x[1] > x[2]) peaks[1] <- TRUE
  if (x[n] >= threshold && x[n] > x[n-1]) peaks[n] <- TRUE
  
  which(peaks)
}

# experimental application with a protein
process_df_peaks <- function(df, threshold = 2) {
  results <- lapply(1:nrow(df), function(i) {
    intensities <- as.numeric(df[i, 2:26])
    protein <- df[i, 1]
    peaks <- find_peaks_diff(intensities, threshold)
    list(Protein = protein, Peaks = peaks)
  })
  
  # save as dataframe
  df_peaks <- data.frame(
    Protein = sapply(results, `[[`, "Protein"),
    Maxima_Positions = sapply(results, function(x) {
      if (length(x$Peaks) == 0) {
        NA_character_
      } else {
        paste(x$Peaks, collapse = ",")
      }
    }),
    stringsAsFactors = FALSE
  )
  
  return(df_peaks)
}

# apply to our dataframe

df_peaks_CTRL <- process_df_peaks(normalized_ctrl_df, threshold = 2)
df_peaks_RNAse <- process_df_peaks(normalized_rnase_df, threshold = 2)


# depict results
head(df_peaks_CTRL)
head(df_peaks_RNAse)

```
Output: all maxima for each protein were identified and localised
protein concentration can be found in multiple fractions 

### global maxima positions
Goal: the indifference in number of maxima between the samples for each protein pose a difficulty for the analysis of the shift
-> evaluate shift based on the global maximum 
```{r}
# Identify global maxima and their position, if present
find_global_maxima <- function(x, threshold = 2) {
  max_val <- max(x)
  if (max_val >= threshold) {
    max_positions <- which(x == max_val)
    if (length(max_positions) == 1) {
      return(max_positions)
    } else {
      middle_index <- mean(range(max_positions))
      return(middle_index)
    }
  } else {
    return(NA_real_)  
  }
}

# extract intensity of the protein concentration in each fraction
process_df_global_maxima <- function(df, threshold = 2) {
  results <- lapply(1:nrow(df), function(i) {
    intensities <- as.numeric(df[i, 2:26])
    protein <- df[i, 1]
    maxima <- find_global_maxima(intensities, threshold)
    list(Protein = protein, Global_Maximum = maxima)
  })
  
  df_maxima <- data.frame(
    Protein = sapply(results, `[[`, "Protein"),
    Global_Maximum = sapply(results, `[[`, "Global_Maximum"),
    stringsAsFactors = FALSE
  )
  
  return(df_maxima)
}

# Apply to our dataframe
global_ctrl <- process_df_global_maxima(normalized_ctrl_df, threshold = 2)
global_rnase <- process_df_global_maxima(normalized_rnase_df, threshold = 2)

head(global_ctrl)
head(global_rnase)

```
# Classification of RBP and Non-RBP according to logistic model
Goal: Apply statistical testscores to our test-model and thus determine the probability of the proteins being RBPs or Non-RBPS
-> logistic model is established using 30 % of the dataset to train the model
-> applying the model to the remaining data yields the predicted probability of RNA-binding for each protein
### shift distance and direction test
```{r}
#Shift Distance and direction based on changes in global maximum
global_ctrl$Global_Maximum <- as.numeric(as.character(global_ctrl$Global_Maximum))
global_rnase$Global_Maximum <- as.numeric(as.character(global_rnase$Global_Maximum))

# Shift distance as absolute value
raw_shift_distance <- global_rnase$Global_Maximum - global_ctrl$Global_Maximum
shift_distance <- abs(raw_shift_distance)

# determine direction
shift_direction <- ifelse(raw_shift_distance > 0, "right shift",
                   ifelse(raw_shift_distance < 0, "left shift",
                          "No Shift"))

shift_result <- data.frame(
  Protein = global_ctrl$Protein,
  Position_Control = global_ctrl$Global_Maximum,
  Position_RNase = global_rnase$Global_Maximum,
  Shift_Distance = shift_distance,    
  Shift_Direction = shift_direction    
)

head(shift_result)
```
- Determining the shift direction based on the position of the global maximum is not ideal, as some RBPs may later lack a clear shift direction.
-> Therefore, this definition of shift direction will not be used.
-> Instead, the shift direction is defined based on the center of mass test

#### skewness of shift distance
Goal: determine skewness of shift distance and check whether robust scaling can be applied
-> if skewness > 1 or < -1, then robust scaling can be applied
robust scaling: (x-median)/IQR -> generally used for asymmetrical distributions
```{r}
#skewness of Shift Distance
# extract values
x <- shift_result$Shift_Distance

# calculate skewness (Fisher-Pearson Skewness)
n <- length(x)
m3 <- sum((x - mean(x, na.rm = TRUE))^3, na.rm = TRUE) / n
s3 <- (sum((x - mean(x, na.rm = TRUE))^2, na.rm = TRUE) / n)^(3/2)
skewness <- m3 / s3
print(paste("Skewness:", round(skewness, 4)))

# plot histogram
hist(x, breaks = 50, col = "lavender", main = "Histogram of shift distance",
     xlab = "shift distance", border = "white")
abline(v = mean(x, na.rm = TRUE), col = "violet", lwd = 2)
text(mean(x, na.rm = TRUE), par("usr")[4]*0.9,
     labels = paste0("Mean = ", round(mean(x, na.rm = TRUE), 2)),
     col = "violet", pos = 4)
```
output: 
- skewness < -1 -> robust scaling can be applied
#### scaling
scaling is needed as different tests are used for the logistic model
-> tests need to be made comparable by scaling them to a common range
```{r}
# Robust scaling function: (x - median) / IQR
robust_scale <- function(x) {
  med <- median(x, na.rm = TRUE)
  iqr <- IQR(x, na.rm = TRUE)
  scaled <- (x - med) / iqr
  return(scaled)
}

# Apply robust scaling to each variable
shift_result$Shift_Distance_scaled <- robust_scale(abs(shift_result$Shift_Distance))

```

```{r}
# scale to 100%
shift_result$Shift_Distance_scaled_100 <- (shift_result$Shift_Distance_scaled/25)*100
```

### change in amplitude of global maximum test
Goal:uncover a possible change in protein-complex assembly after rnase treatment
calculate the amplitude of the global maximum and whether a gain or loss took place 
```{r}
# calculate amplitude of global maximum for control measurements
get_global_amplitudes <- function(expression_df, global_ctrl) {
  amplitudes <- mapply(function(row_idx, max_pos_str) {
    max_positions <- as.numeric(unlist(strsplit(max_pos_str, ",")))
    
    if (length(max_positions) == 0 || all(is.na(max_positions))) {
      return(NA)
    }
    
    intensities <- as.numeric(expression_df[row_idx, max_positions + 1])
    
    paste(intensities, collapse = ",")
  },
  row_idx = 1:nrow(expression_df),
  max_pos_str = as.character(global_ctrl$Global_Maximum)

  )
  
  data.frame(
    Protein = expression_df[[1]],
    Amplitudes = amplitudes,
    stringsAsFactors = FALSE
  )
}
#apply function to our data
global_amplitudes_ctrl <- get_global_amplitudes(normalized_ctrl_df, global_ctrl)

# depict results
head(global_amplitudes_ctrl)

```

```{r}
#calculate amplitude of global maximum for rnase measurments
get_global_amplitudes <- function(expression_df, global_rnase) {
  amplitudes <- mapply(function(row_idx, max_pos_str) {
    max_positions <- as.numeric(unlist(strsplit(max_pos_str, ",")))
    
    if (length(max_positions) == 0 || all(is.na(max_positions))) {
      return(NA)
    }
    
    intensities <- as.numeric(expression_df[row_idx, max_positions + 1])
    
    paste(intensities, collapse = ",")
  },
  row_idx = 1:nrow(expression_df),
  max_pos_str = as.character(global_rnase$Global_Maximum)

  )
  
  data.frame(
    Protein = expression_df[[1]],
    Amplitudes = amplitudes,
    stringsAsFactors = FALSE
  )
}

global_amplitudes_rnase <- get_global_amplitudes(normalized_rnase_df, global_rnase)

head(global_amplitudes_rnase)

```
### gain and loss of amplitude in peak test
```{r}
##Gain and loss of amplitude in peak 
# amplitude as numerics
global_amplitudes_ctrl$Amplitude <- as.numeric(as.character(global_amplitudes_ctrl$Amplitude))
global_amplitudes_rnase$Amplitude <- as.numeric(as.character(global_amplitudes_rnase$Amplitude))

# loss: amplitude is smaller after rnase treatment
amplitude_loss <- pmax(global_amplitudes_ctrl$Amplitude - global_amplitudes_rnase$Amplitude, 0)

# gain: amplitude is greater bevor rnase treatment
amplitude_gain <- pmax(global_amplitudes_rnase$Amplitude - global_amplitudes_ctrl$Amplitude, 0)

amplitude_changes <- data.frame(
  Protein = global_amplitudes_ctrl$Protein,
  Amplitude_Control = global_amplitudes_ctrl$Amplitude,
  Amplitude_RNase = global_amplitudes_rnase$Amplitude,
  loss = amplitude_loss,
  gain = amplitude_gain
)

head(amplitude_changes)
```
Goal: to establish a general shift we only need to assess whether a change in amplitude occurred regardless of the direction
```{r}
###Addition of all amplitude changes
amplitude_changes$changes <- amplitude_changes$loss + amplitude_changes$gain
```
#### skewness of all amplitude changes
```{r}
#skewness of amplitude changes
# extract values
x <- amplitude_changes$changes

# calculate skewness (Fisher-Pearson Skewness)
n <- length(x)
m3 <- sum((x - mean(x, na.rm = TRUE))^3, na.rm = TRUE) / n
s3 <- (sum((x - mean(x, na.rm = TRUE))^2, na.rm = TRUE) / n)^(3/2)
skewness <- m3 / s3
print(paste("Skewness:", round(skewness, 4)))

# plot histogram
hist(x, breaks = 50, col = "lavender", main = "Histogram of amplitude changes",
     xlab = "amplitude changes", border = "white")
abline(v = mean(x, na.rm = TRUE), col = "violet", lwd = 2)
text(mean(x, na.rm = TRUE), par("usr")[4]*0.9,
     labels = paste0("Mean = ", round(mean(x, na.rm = TRUE), 2)),
     col = "violet", pos = 4)
```
output: 
- skewness > 1 -> robust scaling can be applied
#### scaling
```{r}
#robust scaling
amplitude_changes$changes_scaled <- robust_scale(amplitude_changes$changes)

#scale to 100%
amplitude_changes$changes_scaled_100<- (amplitude_changes$changes_scaled/100)*100
head(amplitude_changes$changes_scaled_100)
```

```{r}#
##prepare shift dataframe for linear regression
# save rigth shift proteins as text file
write.table(right_shift_df, file = "Right_Shift_Proteins.txt", sep = "\t", row.names = FALSE, quote = FALSE)

```
### earth mover test
principle of earth mover test: 
- test calculates the work that needs to be done in order to transform one distribution into the other
-> determines mass (amount of protein) that needs to be moved a certain distance in order to compute said transformation
-> for a 1D-EMD the distance between CDFs is determined (distance is positive by definition)
emd <- sum(abs(ecdf(P)(x) - ecdf(Q)(x))) * (max(x) - min(x)) / length(x)
```{r}
install.packages("emdist")  

```

```{r}
library(emdist)

# Custom function to compute 1D Earth Mover's Distance
compute_emd_1d <- function(vec1, vec2) {
  # Remove NAs
  vec1 <- as.numeric(na.omit(vec1))
  vec2 <- as.numeric(na.omit(vec2))
  
  # Ensure both vectors have the same length
  if (length(vec1) != length(vec2)) {
    return(NA)
  }
  
  # Normalize both vectors to sum to 1 (as probability distributions)
  vec1 <- vec1 / sum(vec1)
  vec2 <- vec2 / sum(vec2)
  
  # Compute cumulative distributions
  cdf1 <- cumsum(vec1)
  cdf2 <- cumsum(vec2)
  
  # EMD in 1D is the sum of absolute differences of CDFs
  emd <- sum(abs(cdf1 - cdf2))
  
  return(emd)
}

# Vector to store EMD results
emd_values <- numeric(nrow(normalized_ctrl_df))

# Loop through all proteins (rows)
for (i in 1:nrow(normalized_ctrl_df)) {
  # Extract fraction profiles (columns 2 to 26)
  vec_ctrl <- as.numeric(normalized_ctrl_df[i, 2:26])
  vec_rnase <- as.numeric(normalized_rnase_df[i, 2:26])
  
  # Compute 1D EMD
  emd_values[i] <- compute_emd_1d(vec_ctrl, vec_rnase)
}

# Add EMD results as a new column to the data frame
normalized_ctrl_df$EMD <- emd_values
normalized_rnase_df$EMD <- emd_values  # Optional, for consistency (values are the same as for Ctrl)

#EMD results can potentially range from 0 (Ctrl and RNase are identical) to 25 (Ctrl and RNase are as dissimilar as they can be)

# depict results
head(normalized_ctrl_df[, c(1, 27)])  # Show protein name and EMD

```
#### skewnsess of EMD
```{r}
#skewness of EMD
# extract values
x <- normalized_ctrl_df$EMD

# calculate skewness (Fisher-Pearson Skewness)
n <- length(x)
m3 <- sum((x - mean(x, na.rm = TRUE))^3, na.rm = TRUE) / n
s3 <- (sum((x - mean(x, na.rm = TRUE))^2, na.rm = TRUE) / n)^(3/2)
skewness <- m3 / s3
print(paste("Skewness:", round(skewness, 4)))

# plot histogram
hist(x, breaks = 50, col = "lavender", main = "Histogram of EMD",
     xlab = "EMD value", border = "white")
abline(v = mean(x, na.rm = TRUE), col = "violet", lwd = 2)
text(mean(x, na.rm = TRUE), par("usr")[4]*0.9,
     labels = paste0("Mean = ", round(mean(x, na.rm = TRUE), 2)),
     col = "violet", pos = 4)
```
output:
- skewness > 1 -> robust scaling can be applied
#### scaling 
```{r}
#robust scaling
normalized_ctrl_df$EMD_scaled <- robust_scale(normalized_ctrl_df$EMD)

#scale to 100%
normalized_ctrl_df$EMD_scaled_100 <- (normalized_ctrl_df$EMD_scaled/25)*100
head(normalized_ctrl_df$EMD_scaled_100)
```
### center of mass test
Goal: determine the center of mass of each protein for each treatment (in this case: mass = protein intensity)
com <- sum(i * I_i) / sum(I_i)
i = fractionnumber	
I_i = protein-intensity for the according fraction
```{r}
# center of mass
# function for calculating center of mass
center_of_mass <- function(values) {
  fracs <- seq_along(values)
  if (sum(values, na.rm = TRUE) == 0) return(NA)
  sum(fracs * values, na.rm = TRUE) / sum(values, na.rm = TRUE)
}
# function for calculating maximal possible COM shift
max_possible_com_shift <- function(values) {
  fracs <- seq_along(values)
  if (sum(values, na.rm = TRUE) == 0) return(NA)
}


# remove column with protein names
fractions_control <- normalized_ctrl_df[, 2:26]
fractions_rnase   <- normalized_rnase_df[, 2:26]

# calculate center of mass
com_control <- apply(fractions_control, 1, center_of_mass)
com_rnase   <- apply(fractions_rnase, 1, center_of_mass)

# Delta COM (Difference in Center of Mass)
delta_com <- com_rnase - com_control

# Absolute Delta COM
delta_com_abs <- abs(delta_com)

# Dataframe with results
center_of_mass_df <- data.frame(
  Protein = normalized_ctrl_df[[1]],
  COM_Control = com_control,
  COM_RNase = com_rnase,
  Delta_COM = delta_com,
  Absolute_Delta_COM = delta_com_abs
)

# Output to check results
print(head(center_of_mass_df))
```
## classification: left or right shift
if delta_COM > 0: right shift
if delta_COM < 0: left shift
-> for proteins that are later on classifies as non-shifting this categtory is ignored
```{r}
# define direction of shift
center_of_mass_df$shift_direction <- ifelse(center_of_mass_df$Delta_COM > 0, "right shift",
                    ifelse(center_of_mass_df$Delta_COM < 0, "left shift",
                           "No Shift"))
# show results
head(center_of_mass_df)
```
#### skewness of center of mass
Goal: determine whether the delta center of mass values are skewed or not
If skewed: robust scaling is possible
```{r}
#skewness of center of mass
# extract values
x <- center_of_mass_df$Delta_COM

# calculate skewness (Fisher-Pearson Skewness)
n <- length(x)
m3 <- sum((x - mean(x, na.rm = TRUE))^3, na.rm = TRUE) / n
s3 <- (sum((x - mean(x, na.rm = TRUE))^2, na.rm = TRUE) / n)^(3/2)
skewness <- m3 / s3
print(paste("Skewness:", round(skewness, 4)))

# plot histogram
hist(x, breaks = 50, col = "lavender", main = "Histogram of Delta_COM",
     xlab = "Relative Delta Center of Mass", border = "white")
abline(v = mean(x, na.rm = TRUE), col = "violet", lwd = 2)
text(mean(x, na.rm = TRUE), par("usr")[4]*0.9,
     labels = paste0("Mean = ", round(mean(x, na.rm = TRUE), 2)),
     col = "violet", pos = 4)

```
Output: delta COM is left-skewed
-> scaling is possible
#### scaling
Goal: transform testscore by centering at median and scaling based on the interquartile range
```{r}
#robust scaling
center_of_mass_df$COM_scaled <- robust_scale(center_of_mass_df$Absolute_Delta_COM)

# Normalize scaled values by max fraction delta (24)
center_of_mass_df$COM_scaled_norm <- center_of_mass_df$COM_scaled / 24

# Scale normalized values to percentage (0-100)
center_of_mass_df$COM_scaled_100 <- center_of_mass_df$COM_scaled_norm * 100

# Show first rows of final dataframe
head(center_of_mass_df[, c("Protein", "Delta_COM", "COM_scaled", "COM_scaled_100")])
```
### wilcoxon statistic: position of maxima between replicates of conditions are compared
determine position of global maximum for each replicate of both conditions
```{r}
# determine global maxima from dataframes with still all 3 replicates in them

# prepare data
# remove protein-column
percent_ctrl <- percent_ctrl_df[, -1]
percent_rnase <- percent_rnase_df[, -1]

# define replicates as steps by 3
rep1_cols <- seq(1, ncol(percent_ctrl), by = 3)
rep2_cols <- seq(2, ncol(percent_ctrl), by = 3)
rep3_cols <- seq(3, ncol(percent_ctrl), by = 3)


# formula for extracting global maxima and their position
get_max_and_pos <- function(mat) {
  max_vals <- apply(mat, 1, max, na.rm = TRUE)
  max_pos <- apply(mat, 1, function(x) {
    if (all(is.na(x))) NA else which.max(x)
  })
  list(value = max_vals, pos = max_pos)
}

# formula for extracting position as numeric fraction value
extract_fraction_number <- function(name_vector) {
  as.numeric(sub("Fraction(\\d+)_Rep\\d+", "\\1", name_vector))
}

# apply formulas on both treatments
ctrl_r1 <- get_max_and_pos(percent_ctrl[, rep1_cols])
ctrl_r2 <- get_max_and_pos(percent_ctrl[, rep2_cols])
ctrl_r3 <- get_max_and_pos(percent_ctrl[, rep3_cols])

rnase_r1 <- get_max_and_pos(percent_rnase[, rep1_cols])
rnase_r2 <- get_max_and_pos(percent_rnase[, rep2_cols])
rnase_r3 <- get_max_and_pos(percent_rnase[, rep3_cols])

# define column names
names_r1 <- colnames(percent_ctrl)[rep1_cols]
names_r2 <- colnames(percent_ctrl)[rep2_cols]
names_r3 <- colnames(percent_ctrl)[rep3_cols]

# define protein names
protein_names <- percent_ctrl_df[, 1]

# save maxima results as dataframes
global_max_ctrl_df <- data.frame(
  Protein = protein_names,
  Max_Rep1 = ctrl_r1$value,
  Max_Rep2 = ctrl_r2$value,
  Max_Rep3 = ctrl_r3$value
)

global_max_rnase_df <- data.frame(
  Protein = protein_names,
  Max_Rep1 = rnase_r1$value,
  Max_Rep2 = rnase_r2$value,
  Max_Rep3 = rnase_r3$value
)

# save fraction position as dataframes
fraktion_pos_ctrl_df <- data.frame(
  Protein = protein_names,
  Max_Pos_Rep1 = extract_fraction_number(names_r1[ctrl_r1$pos]),
  Max_Pos_Rep2 = extract_fraction_number(names_r2[ctrl_r2$pos]),
  Max_Pos_Rep3 = extract_fraction_number(names_r3[ctrl_r3$pos])
)

fraktion_pos_rnase_df <- data.frame(
  Protein = protein_names,
  Max_Pos_Rep1 = extract_fraction_number(names_r1[rnase_r1$pos]),
  Max_Pos_Rep2 = extract_fraction_number(names_r2[rnase_r2$pos]),
  Max_Pos_Rep3 = extract_fraction_number(names_r3[rnase_r3$pos])
)

# depict results
print(global_max_ctrl_df)
print(fraktion_pos_ctrl_df)

print(global_max_rnase_df)
print(fraktion_pos_rnase_df)
```

```{r}
#paired Wilxocon test to compare the global maxima positions between control und rnase
# Prepare vectors to store results
p_values <- numeric(nrow(fraktion_pos_ctrl_df))       # p-values
test_statistics <- numeric(nrow(fraktion_pos_ctrl_df))  # test statistic (W)

# Loop through each protein (each row)
for (i in 1:nrow(fraktion_pos_ctrl_df)) {
  # Extract numeric values from columns 2 to 4 for both conditions
  ctrl_values <- as.numeric(fraktion_pos_ctrl_df[i, 2:4])
  rnase_values <- as.numeric(fraktion_pos_rnase_df[i, 2:4])
  
  # Check if all paired values are identical
  if (all(ctrl_values == rnase_values)) {
    # No difference → Wilcoxon test not defined
    p_values[i] <- 1
    test_statistics[i] <- 21  # explanation below
  } else {
    # Run paired Wilcoxon test
    test_result <- wilcox.test(ctrl_values, rnase_values, paired = TRUE, exact = FALSE)
    
    # Store results
    p_values[i] <- test_result$p.value
    test_statistics[i] <- test_result$statistic
  }
}

# Combine results into a dataframe with protein names
Wilcoxon_Pos_Max <- data.frame(
  Protein = fraktion_pos_ctrl_df[[1]],
  P_value = p_values,
  Wilcoxon_statistic = test_statistics
)

# View first few results
head(Wilcoxon_Pos_Max)
```
output:
When all paired differences are zero, the test is undefined, so the statistic is NA.
Using zero or small values for NAs would imply a strong difference (since small W indicates significance), which is incorrect for equal paired samples.
To include these cases in a combined test score where lower W indicates stronger effect, assigning the maximum possible W (e.g., 21 for n=6 pairs) to NAs can be a practical way to treat “no difference” cases as neutral or non-significant.
This avoids interpreting NAs as missing and ensures consistency in scoring, especially if your analysis assumes numeric values for all proteins.
However, this is a heuristic, not a statistical rule, since max W usually indicates a strong positive shift, not equality.
#### skewness of Wilcoxon-statistic
```{r}
#skewness of Wilcoxon-statistic Position of maxima
# extract values
x <- Wilcoxon_Pos_Max$Wilcoxon_statistic

# calculate skewness (Fisher-Pearson Skewness)
n <- length(x)
m3 <- sum((x - mean(x, na.rm = TRUE))^3, na.rm = TRUE) / n
s3 <- (sum((x - mean(x, na.rm = TRUE))^2, na.rm = TRUE) / n)^(3/2)
skewness <- m3 / s3
print(paste("Skewness:", round(skewness, 4)))

# plot histogram
hist(x, breaks = 50, col = "lavender", main = "Wilcoxon-statistic Position of maxima",
     xlab = "Wilcoxon-statistic Position of maxima", border = "white")
abline(v = mean(x, na.rm = TRUE), col = "violet", lwd = 2)
text(mean(x, na.rm = TRUE), par("usr")[4]*0.9,
     labels = paste0("Mean = ", round(mean(x, na.rm = TRUE), 2)),
     col = "violet", pos = 4)
```
output:
- skewness -1 < -0,0508 < 1 -> robust scaling cannot be applied
-> error in Wilcoxon test was recognized too late -> there was no time to find a better suited scaling method for all tests -> for consistency robust scaling was applied nonetheless

#### scaling
```{r}
#robust scaling
Wilcoxon_Pos_Max$Wilcoxon_statistic_scaled<- robust_scale(Wilcoxon_Pos_Max$Wilcoxon_statistic)

#scale to 100%: maximum wilcoxon statistic is sum of ranks (1+2+3+4+5+6 = 21)
Wilcoxon_Pos_Max$Wilcoxon_statistic_scaled_100 <- (Wilcoxon_Pos_Max$Wilcoxon_statistic_scaled/21)*100
head(Wilcoxon_Pos_Max$Wilcoxon_statistic_scaled_100)
```

### test score dataframe
- contains the results of all tests
```{r}
#define dataframe for all test scores of all proteines
test_scores_combined_df <- data.frame(
  Protein = normalized_ctrl_df$Protein,
  Shift_Distance_ctrl_to_rnase = shift_result$Shift_Distance_scaled_100,
  Shift_direction_ctrl_to_rnase = center_of_mass_df$shift_direction,
  amplitude_changes = amplitude_changes$changes_scaled_100,
  EMD = normalized_ctrl_df$EMD_scaled_100,
  Center_of_Mass = center_of_mass_df$COM_scaled_100,
  Wilcoxon_statistic_Max = Wilcoxon_Pos_Max$Wilcoxon_statistic_scaled_100
)
#depict result
  head(test_scores_combined_df)
  
#check that no NAs were created
sum(is.na(test_scores_combined_df))
```


# determine reference proteins
- A dataframe from RDeeP is loaded, which indicates how often each protein has been classified as RNA-binding
- This dataframe includes proteins identified as RNA-binding in any human cell type 

simplifications and limitations: 
- By using this reference, we ignore cell line–specific variation in RNA-binding behavior
- The dataset refers to RNA-binding proteins (RBPs), not necessarily to RNA-dependent proteins (RDPs)
->Therefore, the model can only identify RNA-binding proteins, not all RNA-dependent ones

model setup:
- A logistic regression model is trained using a 30/70 train-test split
- The training set includes 550 positive control proteins (RBPs) and 550 negative control proteins (non-RBPs), which together represent ~30% of all proteins in the dataset
```{r}
# install and load package
install.packages("DT")
library(DT)

# read data
url <- "https://raw.githubusercontent.com/MichelaPozzi/2025-topic-03-group-04/main/table_RBP_lists.csv"
lines <- readLines(url, encoding = "UTF-8")
header <- unlist(strsplit(lines[6], ","))
data <- read.csv(text = lines[7:length(lines)], header = FALSE, sep = ",", stringsAsFactors = FALSE)
colnames(data) <- header

# depict datatable
datatable(data, filter = "top", options = list(pageLength = 25))

# save dataframe locally
write.csv(data, "mein_ergebnis.csv", row.names = FALSE)

#save result as dataframe
RBP_resources <- as.data.frame(data)

#depict result
head(RBP_resources)
```

## negative control
Goal: identify proteins that have never or just once been listed as RBPs
-> increases probability that these proteins do not actually interact with RNA in our cell line
```{r}
#identify proteins which can potentially be used as negative control
# proteins which were identified as RDPs not more than once
RNA_independent <- RBP_resources[RBP_resources[,4] %in% c(0,1), , drop = FALSE]

# depict result
head(RNA_independent)
```
### select 550 porteins as negative control
- compare potential negative control proteins (listed in RDeeP dataframe) to proteins from our datafram 
-> find matches 
-> select 550 of those, ideally only those proteins that have never been caracterized as RBPs 
```{r}
# check which of the potential proteins for negative control are actually contained in our dataset
row_subset <- RNA_independent[,1]
ref_row <- normalized_ctrl_df[,1]

# find common proteins 
common_proteins_negative <- intersect(row_subset, ref_row)

# depict vector with common proteins
length(common_proteins_negative)

# select 550 proteins for negative control
# (to be combined later with 550 positive controls for a balanced dataset of 1100 proteins,
#  which is about 30% of all proteins, suitable for a 30/70 training-test split)

# Filter proteins with '0' in column 4 that are present in the dataset 
zero_proteins <- intersect(RNA_independent[RNA_independent[,4] == 0, 1], normalized_ctrl_df[,1])

# Filter proteins with '1' in column 4 that are present in the dataset 
one_proteins <- intersect(RNA_independent[RNA_independent[,4] == 1, 1], normalized_ctrl_df[,1])

# Count how many definite negatives we have
n_zero <- length(zero_proteins)

# Calculate how many proteins with '1' we need to sample to reach 550 total
n_fill <- 550 - n_zero

# Set random seed for reproducibility
set.seed(42)

# Randomly sample the required number of '1' proteins to fill the negative control set
sampled_one_proteins <- sample(one_proteins, n_fill)

# Combine the definite negatives and sampled proteins to form the final negative control set
negative_control <- c(zero_proteins, sampled_one_proteins)

# Confirm the length of the final negative control vector (should be 550)
length(negative_control)


```

```{r}
# create dataframe negative control
# extract corresponding proteins from test_scores_combined_df
df_negative_control <- test_scores_combined_df[test_scores_combined_df$Protein %in% negative_control, ] 
# depict result
print(df_negative_control)
```

## positive control
Goal: identify proteins that have listed as RBPs most often
```{r}
#identify proteins which can potentially be used as positive control
RNA_dependent <- RBP_resources[RBP_resources[,4] %in% c(8, 9, 10, 11, 12, 13, 14, 15,16, 17, 18), , drop = FALSE]

#depict result
head(RNA_dependent)
```

### select 550 porteins as positive control
- compare potential positive control proteins (listed in RDeeP dataframe) to proteins from our datafram 
-> find matches 
-> select 550 of those
```{r}
#check which of the potential proteins for negative control are actually contained in our dataset
row_subset <- RNA_dependent[,1]
ref_row <- normalized_ctrl_df[,1]

# find common proteins
common_proteins_positive <- intersect(row_subset, ref_row)

# Output the number of common proteins
length(common_proteins_positive)
 
positive_control <- common_proteins_positive
```

```{r}
# create dataframe positive control 
# extract corresponding proteins from test_scores_combined_df 
df_positive_control <- test_scores_combined_df[test_scores_combined_df$Protein %in% positive_control, ] 
# depict result
print(df_positive_control)
dim(df_positive_control)
```

# logistic model
## model setup
Goal: Use the reference proteins to train a logistic model that predicts the probability of RNA binding for the remaining proteins
### comparison of positive and negative control
- A significant difference between positive and negative controls is essential for the model to learn a meaningful separation
- A Mann–Whitney U test is used to assess whether the positive and negative control groups perform significantly differently in the tests
```{r}
#check whether positive and negative control are reliable and whether they differ significantly 
# Vector of column indices to test
cols_to_test <- c(2,4,5,6,7)

# Empty list to store test results
test_results <- list()

for (col in cols_to_test) {
  # Extract the column from both dataframes
  pos_values <- df_positive_control[[col]]
  neg_values <- df_negative_control[[col]]
  
  # Perform Wilcoxon Rank Sum test (Mann-Whitney U test)
  test <- wilcox.test(pos_values, neg_values)
  
  # Store the p-value with the column name as key
  test_results[[colnames(df_positive_control)[col]]] <- test$p.value
}

# Create a data frame to display p-values
test_results_df <- data.frame(
  Column = names(test_results),
  p_value = unlist(test_results)
)

# Print the results
print(test_results_df)
```
output: for all tests there is a significant difference in the performance of positive and negative control proteins
-> reference proteins are suitable for training the logistic model


### visualization of performence of reference proteins in tests
```{r}
# Load libraries
library(ggplot2)
library(dplyr)
library(tidyr)

# Example columns to test (analogous to cols_to_test)
cols_to_test <- c(2,4,5,6,7)
col_names <- colnames(df_positive_control)[cols_to_test]

n <- length(cols_to_test)
x_positions <- seq(1, by = 1, length.out = n)

# Determine range of y-axis values
all_values <- c()
for(i in seq_along(cols_to_test)) {
  col_idx <- cols_to_test[i]
  vals_neg <- as.numeric(df_negative_control[[col_idx]])
  vals_pos <- as.numeric(df_positive_control[[col_idx]])
  all_values <- c(all_values, vals_neg, vals_pos)
}

# Transparent fill colors
fill_neg <- adjustcolor("#c9a0dc", alpha.f = 0.3)
fill_pos <- adjustcolor("#A9C1A6", alpha.f = 0.3)

plot(NA, 
     xlim = c(0.5, n + 0.5), 
     ylim = range(all_values, na.rm = TRUE), 
     xaxt = "n", 
     xlab = "", 
     ylab = "Value", 
     main = "Negative Control vs Positive Control",
     cex.main = 1,
     cex.lab = 1
)

# Draw x-axis without labels
axis(1, at = x_positions, labels = FALSE, tick = TRUE)

# Draw rotated labels below the axis
text(x = x_positions, y = par("usr")[3] - 0.05 * diff(par("usr")[3:4]), 
     labels = col_names, 
     srt = 45,          # 45 degree angle
     adj = 1,           # right aligned
     xpd = TRUE,        # allow drawing outside plot region
     cex = 0.5)         # small font size

for(i in seq_along(cols_to_test)) {
  col_idx <- cols_to_test[i]
  x <- x_positions[i]
  
  vals_neg <- as.numeric(df_negative_control[[col_idx]])
  vals_pos <- as.numeric(df_positive_control[[col_idx]])
  
  if(sum(!is.na(vals_neg)) < 2 || sum(!is.na(vals_pos)) < 2) next
  
  dens_neg <- tryCatch(density(vals_neg, na.rm = TRUE), error = function(e) NULL)
  dens_pos <- tryCatch(density(vals_pos, na.rm = TRUE), error = function(e) NULL)
  if(is.null(dens_neg) || is.null(dens_pos)) next
  
  max_y <- max(c(dens_neg$y, dens_pos$y))
  scale_factor <- 0.4 / max_y
  
  # Left half (negative control) with transparent fill
  x_left <- x - dens_neg$y * scale_factor
  y_vals <- dens_neg$x
  polygon(c(x_left, rev(rep(x, length(x_left)))), 
          c(y_vals, rev(y_vals)), 
          col = fill_neg, border = "#c9a0dc")
  
  # Right half (positive control) with transparent fill
  x_right <- x + dens_pos$y * scale_factor
  y_vals2 <- dens_pos$x
  polygon(c(rep(x, length(x_right)), rev(x_right)), 
          c(y_vals2, rev(y_vals2)), 
          col = fill_pos, border = "#A9C1A6")
  
  # Median lines
  med_neg <- median(vals_neg, na.rm = TRUE)
  med_pos <- median(vals_pos, na.rm = TRUE)
  
  lines(c(x - 0.2, x), rep(med_neg, 2), col = "#c9a0dc", lty = 2, lwd = 2)
  lines(c(x, x + 0.2), rep(med_pos, 2), col = "#A9C1A6", lty = 2, lwd = 2)
}

legend("topright",
       legend = c("Negative Control", "Positive Control"),
       fill = c(fill_neg, fill_pos),    # Transparent fill colors in legend
       border = c("#c9a0dc", "#A9C1A6"),
       bty = "n",
       cex = 0.8)

```


### build logistic model
- logistic model based on all tests is created
- check for redundancy/multicollinearity using step()function
```{r}
# 1. Add labels to the datasets
df_positive_control$label <- 1  # Positive control group labeled as 1
df_negative_control$label <- 0  # Negative control group labeled as 0

# 2. Combine the datasets and keep the relevant columns (features + label)
df_all <- rbind(
  df_positive_control[, c("EMD", "Shift_Distance_ctrl_to_rnase", "amplitude_changes", "Center_of_Mass", "Wilcoxon_statistic_Max", "label")],
  df_negative_control[, c("EMD", "Shift_Distance_ctrl_to_rnase", "amplitude_changes", "Center_of_Mass", "Wilcoxon_statistic_Max", "label")]
)

# 3. Fit full logistic regression model with all predictors
full_model <- glm(label ~ EMD + Shift_Distance_ctrl_to_rnase + amplitude_changes + Center_of_Mass + Wilcoxon_statistic_Max,
                  data = df_all,
                  family = binomial)

# 4. Stepwise feature selection based on AIC
step_model <- step(full_model, direction = "both", trace = TRUE)

# 5. Summary of the reduced model
summary(step_model)

# 6. Create a weighted score using the selected model's coefficients
coeffs <- coef(step_model)[-1]  # exclude intercept
selected_features <- names(coeffs)
df_all$weighted_score <- rowSums(sweep(df_all[, selected_features], 2, coeffs, FUN = "*"))

# 7. Visualize weighted score by class
boxplot(weighted_score ~ label, data = df_all,
        names = c("Negative", "Positive"),
        xlab = "Class Label", ylab = "Weighted Score",
        main = "Distribution of Weighted Scores by Class")

```
output: 
- step()functionn did not remove any predictors -> no redundancy problems
- The regression estimate represents the influence of a given predictor on the log-odds of RNA binding. A higher estimate increases the predicted probability.
- The Pr(>|z|) value indicates whether a test is statistically significant, the lower values (e.g. p<0.05), the more significant the predictor’s contribution to the model (+ the less likely is the relation coincidental)

## visulaization of the tests' influence on the final probability score
```{r}
logistic_model <- glm(label ~ EMD + Shift_Distance_ctrl_to_rnase + amplitude_changes + Center_of_Mass + Wilcoxon_statistic_Max,
                      data = df_all,
                      family = "binomial")


# Check if the model is a logistic regression (binomial family)
if (inherits(logistic_model, "glm") && family(logistic_model)$family == "binomial") {
  
  # Extract coefficients (excluding intercept)
  coeffs <- coef(logistic_model)
  coeffs <- coeffs[-1]  # Remove intercept
  
  # Create a data frame for plotting
  df_coeffs <- data.frame(
    Feature = names(coeffs),
    Coefficient = coeffs,
    Importance = abs(coeffs)
  )
  
  # Optional: sort by absolute importance descending
  df_coeffs <- df_coeffs[order(-df_coeffs$Importance), ]
  
  # Visualization using ggplot2
  library(ggplot2)
  ggplot(df_coeffs, aes(x = reorder(Feature, Importance), y = Coefficient, fill = Coefficient > 0)) +
    geom_col() +
    coord_flip() +  # horizontal bars
    scale_fill_manual(values = c("TRUE" = "#c9a0dc", "FALSE" = "#A9C1A6"), guide = "none") +
    labs(
      subtitle = "Feature Importance in Logistic Regression Model",
      x = "Feature",
      y = "Regression Coefficient"
    ) +
    theme_minimal(base_size = 14)
  
} else {
  stop("The model is not a logistic regression with binomial family.")
}

```
output:
- positive coefficients (purple bars) indicate features that increase the probability of RNA-binding
- negative coefficients (green bars) indicate features that decrease the probability of RNA-binding
-> EMD has the strongest influence on the probability of RNA-binding

## ROC Curve Visualizing test-model performance
Goal: evaluate the predictive accuracy of the logistic regression model by showing the trade-off between the true positive rate (sensitivity) and false positive rate (1 - specificity) across different classification thresholds
```{r}
# Predicted probabilities from the model (as before)
predicted_probs <- predict(logistic_model, type = "response")

# True labels
labels <- df_all$label

# Thresholds from 0 to 1 in small steps
thresholds <- seq(0, 1, by = 0.01)

# Initialize vectors for True Positive Rate (TPR) and False Positive Rate (FPR)
tpr <- numeric(length(thresholds))
fpr <- numeric(length(thresholds))

for (i in seq_along(thresholds)) {
  thresh <- thresholds[i]
  
  # Predictions based on the threshold
  preds <- ifelse(predicted_probs >= thresh, 1, 0)
  
  # Calculate True Positives, False Positives, True Negatives, False Negatives
  TP <- sum(preds == 1 & labels == 1)
  FP <- sum(preds == 1 & labels == 0)
  TN <- sum(preds == 0 & labels == 0)
  FN <- sum(preds == 0 & labels == 1)
  
  # Calculate TPR and FPR
  tpr[i] <- ifelse((TP + FN) == 0, 0, TP / (TP + FN))
  fpr[i] <- ifelse((FP + TN) == 0, 0, FP / (FP + TN))
}

# Define colors: lavender for ROC curve, beige for diagonal line
lavender <- "#c9a0dc"
graygreen <- "#A9C1A6"

# Plot ROC curve with lavender line
plot(fpr, tpr, type = "l", col = lavender, lwd = 2,
     xlab = "False Positive Rate (FPR)", ylab = "True Positive Rate (TPR)",
     main = "ROC Curve")

# Add diagonal line (random guess) in beige
abline(a = 0, b = 1, lty = 2, col = graygreen)
```


### test-model applied to testscores of remaining proteins
Goal: predict probability of RNA binding proteins
```{r}
#apply model to our data
# 1. Calculate predicted probabilities of RNA-binding using the logistic model
test_scores_combined_df$probability <- predict(logistic_model, newdata = test_scores_combined_df, type = "response")

# 2. View the first few rows of the result
head(test_scores_combined_df)
```
#### threshold
Goal: define threshold with a 10% FDR
```{r}
#calculate threshold automatically
# predictions
test_scores_combined_df$predicted_prob <- predict(logistic_model, newdata = test_scores_combined_df, type = "response")

# Marks for reference
test_scores_combined_df$is_negative <- test_scores_combined_df$Protein %in% df_negative_control$Protein

# Evaluate thresholds 
thresholds <- seq(0, 1, by = 0.001)
fdr_values <- sapply(thresholds, function(thresh) {
  selected <- test_scores_combined_df[test_scores_combined_df$predicted_prob > thresh, ]
  if (nrow(selected) == 0) return(NA)
  
  false_positives <- sum(selected$is_negative, na.rm = TRUE)
  all_positives   <- nrow(selected)
  fdr <- false_positives / all_positives
  return(fdr)
})

# Find threshold at FDR ≤ 10% 
target_fdr <- 0.10
valid_thresholds <- thresholds[which(fdr_values <= target_fdr)]
optimal_threshold <- if (length(valid_thresholds) > 0) min(valid_thresholds) else NA

cat("Threshold bei FDR ≤ 10%:", optimal_threshold, "\n")

#visualize FDR
plot(thresholds, fdr_values, type = "l", col = "lavender", lwd = 2,
     xlab = "Prediction Threshold", ylab = "FDR",
     main = "FDR vs Threshold")
abline(h = 0.1, col = "violet", lty = 2)
abline(v = optimal_threshold, col = "lightgreen", lty = 2)

# Proteins with prediction scores above the optimal threshold
significant_shifted_proteins <- test_scores_combined_df[test_scores_combined_df$probability > optimal_threshold, ]

# Number of significantly shifting proteins
num_significant_shifted <- nrow(significant_shifted_proteins)

cat("Number of significantly shifted proteins:", num_significant_shifted, "\n")
```
### overview of RBPs
- Evaluate the composition of proteins classified as RNA-binding by the logistic model:
-> How many belong to the positive control set?
-> many are false positives (i.e., negative control proteins)?
-> How many are novel predictions outside of the control sets?
```{r}
#threshold is selected at FDR = 10%
# 1. Mark reference proteins in the full dataset
test_scores_combined_df$is_reference <- test_scores_combined_df$Protein %in% df_positive_control$Protein

# 2. Make predictions using the logistic model
predictions <- predict(logistic_model, newdata = test_scores_combined_df, type = "response")

# 3. Apply threshold (e.g., 0.95) to get predicted positives
selected_proteins <- test_scores_combined_df[predictions > optimal_threshold, ]
selected_proteins$is_reference <- selected_proteins$Protein %in% df_positive_control$Protein


# 4. Count how many of the selected proteins are reference proteins
ref_counts <- table(selected_proteins$is_reference)
names(ref_counts) <- c("Not positive control protein (FALSE)", "Positive Control protein (TRUE)")
print(ref_counts)

#how many proteins of negative control are found at this threshold
# 1. Mark negative control proteins in the full dataset
test_scores_combined_df$is_negative <- test_scores_combined_df$Protein %in% df_negative_control$Protein

# 2. Make predictions using the existing logistic model
predictions <- predict(logistic_model, newdata = test_scores_combined_df, type = "response")

# 3. Apply threshold (z.B. gleich wie oben: 0.3899612)
selected_proteins <- test_scores_combined_df[predictions > optimal_threshold, ]

# 4. Count how many of the selected proteins are from the negative control
neg_counts <- table(selected_proteins$is_negative)
names(neg_counts) <- c("Not negative control protein (FALSE)", "Negative control protein (TRUE)")
print(neg_counts)

# Extract those negative control proteins among predicted positives
predicted_negatives <- selected_proteins[selected_proteins$is_negative, ]

# Extract predicted hits that are NOT from the negative control -> TRUE: number of proteins from negative-control that are classified as shifting from the model
predicted_non_negative <- selected_proteins[!selected_proteins$is_negative, ]
```
### right shift proteins
Goal: define right shift proteins according to their delta COM
```{r}
# filter proteins for significant right shift with defined threshold
filtered_df <- test_scores_combined_df[
  test_scores_combined_df$Shift_direction_ctrl_to_rnase == "right shift" &
  test_scores_combined_df$probability >= optimal_threshold, 
]

# Extract protein names from first column, save as dataframe
right_shift_proteins_df <- data.frame(Protein = filtered_df[[1]])

# Number of filtered proteins
right_shift_count <- nrow(filtered_df)

# Depict results
print(right_shift_proteins_df)
print(paste("Anzahl der 'right shift'-Proteine mit Probability ≥", optimal_threshold, ":", right_shift_count))
```

# visualisation of test-model results 
## composition of found RBPS
```{r}
# DataFrames mit Proteinnamen
total_proteins <- data.frame(Protein = test_scores_combined_df$Protein)
num_rbp <- data.frame(Protein = df_positive_control$Protein)

# Anzahl berechnen
total_count <- nrow(total_proteins)
rbp_count <- nrow(num_rbp)

# Farben
colors1 <- c("#F5E1A4","#c9a0dc")

# Matrix erstellen für Barplot
counts_matrix1 <- matrix(c(
  total_count - rbp_count,  
  rbp_count
), nrow = 2, byrow = TRUE)

colnames(counts_matrix1) <- c("All Proteins")
rownames(counts_matrix1) <- c("Non-shifted", "Shifted")

# Barplot
barplot(counts_matrix1,
        beside = FALSE,
        col = colors1,
        border = "white",
        main = "All Proteins vs. Significant Shifts",
        ylab = "Number of Proteins",
        las = 1)

# Legende
legend("topright",
       legend = rownames(counts_matrix1),
       fill = colors1,
       bty = "n")


# Define colors for next plot (reversed order)
colors2 <- c("#c9a0dc", "#F5E1A4")

# 2. Plot: Direction of Protein Shifts (Left vs. Right)

# define number of proteins
num_right_shift <- right_shift_count
num_left_shift <- num_significant_shifted - right_shift_count

counts_matrix2 <- matrix(c(
  num_right_shift,  # Number of proteins shifted to the right
  num_left_shift    # Number of proteins shifted to the left
), nrow = 2, byrow = TRUE)

colnames(counts_matrix2) <- c("Shifted Proteins")
rownames(counts_matrix2) <- c("Right Shift", "Left Shift")

barplot(counts_matrix2,
        beside = FALSE,
        col = colors2,
        border = "white",
        main = "Direction of Protein Shifts",
        ylab = "Number of Proteins",
        las = 1,
        ylim = c(0, num_significant_shifted))

legend("topright",
       legend = rownames(counts_matrix2),
       fill = colors2,
       bty = "n")

# Colors: True Positives purple, False Positives beige, Newly Identified RBPs gray-green
colors3 <- c("Beige" = "#F5E1A4", "True Positives" = "#c9a0dc", "New RBPs" = "#A9C1A6")

# Number of each group
false_positives_num <- sum(predicted_negatives == "TRUE")  # Number of False Positives
true_positives_num <- sum(predicted_non_negative == "TRUE") # Number of True Positives

# Total number of predicted proteins (example)
total_predicted_proteins <- num_significant_shifted  # Or another appropriate number

# Number of newly identified RBPs
new_rbps_num <- total_predicted_proteins - (false_positives_num + true_positives_num)
if (new_rbps_num < 0) new_rbps_num <- 0  # Safety check

# Matrix for barplot: order = False Positives, True Positives, New RBPs
counts_matrix3 <- matrix(c(false_positives_num, true_positives_num, new_rbps_num),
                         nrow = 3, byrow = TRUE)

colnames(counts_matrix3) <- c("Prediction Quality")
rownames(counts_matrix3) <- c("False Positives", "True Positives", "New RBPs")

# Create barplot
barplot(counts_matrix3,
        beside = FALSE,
        col = colors3,
        border = "white",
        main = "Prediction Quality with Newly Identified RBPs",
        ylab = "Number of Proteins",
        las = 1,
        ylim = c(0, total_predicted_proteins))

# Add legend
legend("topright",
       legend = rownames(counts_matrix3),
       fill = colors3,
       bty = "n")
```

### visualization of shift direction
```{r}
# Join dataframes including global maxima and calculate their differences
df_diff_peaks <- center_of_mass_df[, 2:6]

df_diff_peaks$diff <- center_of_mass_df$Delta_COM

# calculate limits for axes
max_val <- max(
  df_diff_peaks$COM_Control,
  df_diff_peaks$COM_RNase_Control,
  na.rm = TRUE
) * 1.1

colors <- ifelse(df_diff_peaks$diff > 0 & test_scores_combined_df$probability > optimal_threshold, "#c9a0dc",      # right shift 
                 ifelse(df_diff_peaks$diff < 0 & test_scores_combined_df$probability > optimal_threshold, "#A9C1A6", # left shift
                        "#F5E1A4"))   # no shift
                  



# Set layout for plot and separate legend area
layout(matrix(1:2, nrow = 1), widths = c(4, 1))

# Plot margins
par(mar = c(5, 4, 4, 1))  # Bottom, left, top, right

# Plot the scatterplot
plot(
  df_diff_peaks$COM_RNase,
  df_diff_peaks$COM_Control,
  xlim = c(0, max_val),
  ylim = c(0, max_val),
  xlab = "Center of Mass RNase",
  ylab = "Center of Mass Control",
  main = "Shift Direction",
  pch = 19,
  col = colors
)

# Reference line y = x
abline(a = 0, b = 1, col = "grey", lty = 2)

# Switch to legend area (right panel)
par(mar = c(0, 0, 4, 0))  # No axis, top margin for title

# Draw an empty plot for the legend
plot.new()

# Draw well-formatted legend with clearer colors
legend("center",
       legend = c(
         expression("Right shift ("*Delta*"COM > 0)"),
         expression("Left shift ("*Delta*"COM < 0)"),
         expression("No shift ("*Delta*"COM = 0)")
       ),
       col = c("#c9a0dc", "#A9C1A6", "#F5E1A4"),
       pch = 19,
       title = "Shift Direction",
       bty = "n",
       pt.cex = 1,     
       text.col = "black",
       cex = 0.7       
)

```


## show result for any desired protein
Enter a protein name to visualize its distribution across conditions (Control and RNase).
The code will also return the predicted probability of RNA-binding and the corresponding classification as either RBP or Non-RBP
```{r}
# Define protein name
protein_name <- "1433B_HUMAN"

# Check if protein exists in both datasets (columns 1 to 26 only)
subset_ctrl <- normalized_ctrl_df[,1:26]
subset_rnase <- normalized_rnase_df[,1:26]

if (!(protein_name %in% subset_ctrl$Protein) |
    !(protein_name %in% subset_rnase$Protein)) {
  stop("Protein not found in one or both datasets (rows 1 to 26).")
}

# Extract only data for the given protein from rows 1–26
ctrl_data <- subset_ctrl[subset_ctrl$Protein == protein_name, ]
rnase_data <- subset_rnase[subset_rnase$Protein == protein_name, ]

# Add condition labels
ctrl_data$Condition <- "Control"
rnase_data$Condition <- "RNase"

# Combine both into one data frame for plotting
combined_df <- rbind(ctrl_data, rnase_data)

# Get probability and RBP classification
prob <- test_scores_combined_df$probability[test_scores_combined_df$Protein == protein_name]
is_rbp <- ifelse(prob > optimal_threshold, "RBP", "Non-RBP")

library(tidyr)
library(ggplot2)
library(dplyr)

# Pivot from wide to long format (gather Fraction1 to Fraction26)
long_df <- pivot_longer(
  combined_df,
  cols = starts_with("Fraction"),
  names_to = "Fraction",
  values_to = "Intensity"
)

# Convert Fraction names to numeric (e.g., "Fraction1" -> 1)
long_df$Fraction <- as.numeric(gsub("Fraction", "", long_df$Fraction))

# Get predicted probability and classification
prob <- test_scores_combined_df$probability[test_scores_combined_df$Protein == protein_name]
is_rbp <- ifelse(prob > optimal_threshold, "RBP", "Non-RBP")

#Plot
ggplot(long_df, aes(x = Fraction, y = Intensity, color = Condition, fill = Condition)) +
  # Fläche unter der Kurve einfärben (aber nicht in der Legende anzeigen)
  geom_area(alpha = 0.3, position = "identity", show.legend = FALSE) +
  # Linie zeichnen (erscheint in der Legende)
  geom_line(linewidth = 1) +
  # Farben manuell setzen
  scale_color_manual(values = c("Control" = "#c9a0dc", "RNase" = "#A9C1A6")) +
  scale_fill_manual(values = c("Control" = "#c9a0dc", "RNase" = "#A9C1A6")) +
  labs(
    title = paste("Profile for Protein:", protein_name),
    subtitle = paste0("Predicted Probability: ", round(prob, 3),
                      " | Classified as: ", is_rbp),
    x = "Fraction",
    y = "Normalized Amount Of Protein",
    color = "Condition"  # Legend title
  ) +
  theme_minimal()

```

```{r}#
install.packages("shiny")
```
# 'app' panel: Protein Profile Viewer
more userfriendly way to depict results: distributions, probability and classification
be mindful: this chunk has to be closed or newly opened window has to be closed for following chunks to run
```{r}
# load libraries
library(shiny)
library(ggplot2)
library(dplyr)
library(tidyr)

# Define User Interface
ui <- fluidPage(
  titlePanel("Protein Profile Viewer"),
  sidebarLayout(
    sidebarPanel(
      textInput("protein_name", "Enter Protein Name:", ""),
      actionButton("plot_button", "Plot")
    ),
    mainPanel(
      plotOutput("proteinPlot")
    )
  )
)

# Define Server logic
server <- function(input, output, session) {
  
  observeEvent(input$plot_button, {
    protein_name <- input$protein_name
    
    if (!exists("normalized_ctrl_df") || !exists("normalized_rnase_df") || 
    !exists("test_scores_combined_df") || !exists("optimal_threshold")) {
  output$proteinPlot <- renderPlot(NULL)
  showNotification("One or more required data objects are missing.", type = "error")
  return()
}
    
    # Select columns 1 (name) and 2:26 (Fraction1 to Fraction25)
    ctrl <- normalized_ctrl_df[, c(1, 2:26)]
    rnase <- normalized_rnase_df[, c(1, 2:26)]
    
    # Filter for selected protein
    ctrl_data <- ctrl %>% filter(Protein == protein_name)
    rnase_data <- rnase %>% filter(Protein == protein_name)
    
    # Check if required data objects are available
    if (!exists("normalized_ctrl_df") || 
        !exists("normalized_rnase_df") || 
        !exists("test_scores_combined_df") || 
        !exists("optimal_threshold")) {
      
      output$proteinPlot <- renderPlot(NULL)
      showNotification("One or more required data objects are missing.", type = "error")
      return()
    }
    
    # Label condition
    ctrl_data$Condition <- "Control"
    rnase_data$Condition <- "RNase"
    
    # Combine and reshape
    combined <- bind_rows(ctrl_data, rnase_data)
    long_df <- combined %>%
      pivot_longer(cols = starts_with("Fraction"),
                   names_to = "Fraction", values_to = "Intensity") %>%
      mutate(Fraction = as.numeric(gsub("Fraction", "", Fraction)))
    
    # Get model probability and RBP classification
    prob <- test_scores_combined_df$probability[test_scores_combined_df$Protein == protein_name]
    if (length(prob) == 0) prob <- NA
    classification <- ifelse(!is.na(prob) & prob > optimal_threshold, "RBP", "Non-RBP")
    
    output$proteinPlot <- renderPlot({
      ggplot(long_df, aes(x = Fraction, y = Intensity, color = Condition, fill = Condition)) +
        geom_area(alpha = 0.3, position = "identity", show.legend = FALSE) +
        geom_line(linewidth = 1) +
        scale_color_manual(values = c("Control" = "#c9a0dc", "RNase" = "#A9C1A6")) +
        scale_fill_manual(values = c("Control" = "#c9a0dc", "RNase" = "#A9C1A6")) +
        scale_x_continuous(breaks = seq(5, 25, by = 5), limits = c(1, 25)) +
        labs(
          title = paste("Profile for Protein:", protein_name),
          subtitle = paste0("Predicted Probability: ", round(prob, 3),
                            " | Classified as: ", classification),
          x = "Fraction",
          y = "Normalized Intensity",
          color = "Condition"
        ) +
        theme_minimal() +
        theme(panel.grid.major.x = element_blank(),
              panel.grid.minor.x = element_blank())
    })
  })
}

shinyApp(ui, server)





```

## optional - save right shift proteins as text file
```{r}#
# Nur die Protein-Namen als Vektor
protein_names <- right_shift_proteins_df$Protein

# Schreibe sie zeilenweise in eine Textdatei
writeLines(protein_names, "right_shift_proteins.txt")
```


#### violinplots
Goal: show the distribution of the protein in direct comparison 
```{r}
# visualize the shift/non-shift of three selected proteins as violinplot
proteins <- c("RLA1_HUMAN", "APC16_HUMAN", "CH033_HUMAN")

# Plotpreperation
n <- length(proteins)
x_positions <- seq(1, by = 1, length.out = n)

# Y-axes
all_values <- c()
for (p in proteins) {
  v1 <- as.numeric(normalized_ctrl_df[normalized_ctrl_df[[1]] == p, 2:26])
  v2 <- as.numeric(normalized_rnase_df[normalized_rnase_df[[1]] == p, 2:26])
  all_values <- c(all_values, v1, v2)
}

# Basisplot
plot(NA,
     xlim = c(0.5, n + 0.5),
     ylim = range(all_values, na.rm = TRUE),
     xaxt = "n",
     xlab = "",
     ylab = "Intensity",
     main = "Symmetric Violinplots")

axis(1, at = x_positions, labels = proteins, las = 2)

# plot
for (i in seq_along(proteins)) {
  protein <- proteins[i]
  x <- x_positions[i]

  ctrl_vals <- as.numeric(normalized_ctrl_df[normalized_ctrl_df[[1]] == protein, 2:26])
  rnase_vals <- as.numeric(normalized_rnase_df[normalized_rnase_df[[1]] == protein, 2:26])
  
  
  if (sum(!is.na(ctrl_vals)) < 2 || sum(!is.na(rnase_vals)) < 2) {
    next
  }

  # calculate density and check validity
  dens_ctrl <- tryCatch(density(ctrl_vals, na.rm = TRUE), error = function(e) NULL)
  dens_rnase <- tryCatch(density(rnase_vals, na.rm = TRUE), error = function(e) NULL)
  
  if (is.null(dens_ctrl) || is.null(dens_rnase)) {
    next
  }
  
  # check length of both x- and y-vector -> have to be the same length
  if (length(dens_ctrl$x) != length(dens_ctrl$y) || length(dens_rnase$x) != length(dens_rnase$y)) {
    next
  }

  # scaling
  max_y <- max(c(dens_ctrl$y, dens_rnase$y))
  scale_factor <- 0.4 / max_y

  # Violine: Control on left
x_left  <- x - dens_ctrl$y * scale_factor
y_vals  <- dens_ctrl$x
polygon(
  x = c(x_left, rev(x_left)),
  y = c(y_vals, rev(y_vals)),
  col = "lightblue", border = "blue"
)

# Violine: RNase on right
x_right <- x + dens_rnase$y * scale_factor
y_vals2 <- dens_rnase$x
polygon(
  x = c(x_right, rev(x_right)),
  y = c(y_vals2, rev(y_vals2)),
  col = "lightpink", border="red"
)

  # add median line
  med_ctrl <- median(ctrl_vals, na.rm = TRUE)
  med_rnase <- median(rnase_vals, na.rm = TRUE)

  lines(c(x - 0.2, x), rep(med_ctrl, 2), col = "blue", lty = 2)
  lines(c(x, x + 0.2), rep(med_rnase, 2), col = "red", lty = 2)
}

# add legend
legend("topright",
       legend = c("Control", "RNase"),
       fill = c("lightblue", "lightpink"),
       border = c("blue", "red"),
       bty="n")
```
Output: visualization of a negative control and positive control protein, as well as one newly identified as shifting

# 4. Detection of features predictive of RNA-binding
# linear regression
Goal: Identify experimentally accessible predictors of RNA-binding
-> establish relation between predictors and RNA-binding behavior

### connect information on protein features to proteins from dataset
### load data with protein features 
-> dataframe from RBP2GO is loaded that contains information on pI, Mass in kDa and length in AA as possible predictors for RNA-binding
```{r}
### read Table_RBPs

# read table
raw <- readLines("data/Table_HS_RBP.txt")
# extract everything below row 7
raw_data <- raw[7:length(raw)]

# clean the table
clean_data <- gsub('\\"', '', raw_data)
writeLines(clean_data, "cleaned_table.txt")
Table_HS_RBP = read.delim("cleaned_table.txt",
                    sep = "\t",
                    header = TRUE,
                    stringsAsFactors = FALSE,
                    check.names = FALSE)
head(Table_HS_RBP)

```

```{r}
# read Table_Non_RBP


# whole table
raw <- readLines("data/Table_HS_Non_RBP.txt")
# extract everything below row 7
raw_data <- raw[7:length(raw)]

# clean the table 
clean_data <- gsub('\\"', '', raw_data)
writeLines(clean_data, "cleaned_table.txt")
Table_HS_Non_RBP = read.delim("cleaned_table.txt",
                    sep = "\t",
                    header = TRUE,
                    stringsAsFactors = FALSE,
                    check.names = FALSE)
head(Table_HS_Non_RBP)

```

```{r}
# check if all our proteins are included in the tables 
all_proteins <- test_scores_combined_df$Protein
proteins_rbp <- Table_HS_RBP$Entry_Name
proteins_non_rbp <- Table_HS_Non_RBP$Entry_Name

is_in_rbp_or_nonrbp <- test_scores_combined_df %in% proteins_rbp | test_scores_combined_df %in% proteins_non_rbp

all(is_in_rbp_or_nonrbp)

```


```{r}
# Select only relevant columns from the RBP and Non-RBP tables
rbp_subset <- Table_HS_RBP[, c("Entry_Name", "pI", "Mass_kDa" , "Length_AA")]
non_rbp_subset <- Table_HS_Non_RBP[, c("Entry_Name", "pI", "Mass_kDa", "Length_AA")]

# Combine both subsets into a single table
Table_HS_Combined <- rbind(rbp_subset, non_rbp_subset)

# Merge: Proteinname-Spalten heißen unterschiedlich -> by.x / by.y verwenden
merged_df <- merge(test_scores_combined_df, Table_HS_Combined,
                   by.x = "Protein", by.y = "Entry_Name",
                   all.x = TRUE)  # keep all entries from test_scores_combined_df
#depict result
head(merged_df)
dim(merged_df)
```

## determinantion of suitable predictors for multiple regression
- Independent variables should exhibit minimal correlation to avoid multicollinearity
-> correlation between molecular weight, sequence length, and isoelectric point was examined
```{r}
# select relevant columns
df_complete <- merged_df[complete.cases(merged_df[, c("probability", "pI", "Mass_kDa")]), ]
# which parameters can be used in multiple regression: correlation between independent variables should be as low as possible

cor(df_complete$Mass_kDa, df_complete$pI, method = "pearson")
cor(df_complete$Mass_kDa, df_complete$Length_AA, method = "pearson")
cor(df_complete$pI, df_complete$Length_AA, method = "pearson")

```
output:
- there is no strong correlation between pI and mass nor between pI and length
- a strong correlation was observed between molecular weight and sequence length
-> since molecular weight is more readily accessible experimentally than sequence length, it was used as a predictor in the multiple regression model along with the isoelectric point (pI)


## multiple regression
goal: assess whether isoelectric point (pI) and molecular mass are reliable predictors of RNA-binding probability
```{r}
# multiple regression: probability ~ pI + Mass_kDa
model <- lm(probability ~ pI + Mass_kDa, data = df_complete)

summary(model)

```
output:
- both the isoelectric point (pI) and molecular weight (Mass_kDa) are significantly associated with the target variable (p < 0.001)
- pI has the stronger effect (t = 20.70), followed by molecular weight (t = 3.38) 
- After accounting for mass, proteins with higher pI tend to have a higher probability of RNA-binding, and similarly, after accounting for pI, higher mass is associated with higher probability
- the model explains approximately 10.8% of the variance (adjusted R² = 0.108)
- model with two predictors is significantly better than null model

### visualization of multiple regression
- hyperplane depicts relation between probability and predictors according to the multiple regression model
```{r}

# visualization of linear regression using hyperplanes

install.packages("scatterplot3d") 
library(scatterplot3d)


depth <- df_complete$probability

# colorscheme
farben <- colorRampPalette(c("#b19cd9", "#7b59b9", "#4b0082"))(100)

color_index <- cut(depth, breaks = 100, labels = FALSE)
point_colors <- farben[color_index]

s3d <- scatterplot3d(df_complete$pI, df_complete$Mass_kDa, df_complete$probability,
                     xlab = "pI", ylab = "Masse (kDa)", zlab = "Probability",
                     pch = 16, color = point_colors,
                     highlight.3d = FALSE,
                     type = "p", angle = 55)

s3d$plane3d(model, col = "#4b0082")



# Koordinaten der Regressionsfläche berechnen
plane <- predict(model, 
                 newdata = data.frame(pI = df_complete$pI, 
                                      Mass_kDa = df_complete$Mass_kDa))

# Fläche einzeichnen
s3d$plane3d(model)
```
output:
- The closer the points lie to the plane, the better the model fits the observed data
- Large distances between the points and the plane indicate that the model explains only a small portion of the variance
-> depicts that only approximately 11% of variance is explained by the model

### visualization of multiple regression
interactive hyperplane displays the same relatins as the chunk above but offers a better spacial unterstanding
```{r}
# create interactive hyperplane 


install.packages("plotly")
library(plotly)

# new classification
df_complete$RBP_shift_category <- ifelse(df_complete$probability >= optimal_threshold,
                                         ifelse(df_complete$Shift_direction_ctrl_to_rnase == "left shift", "RBP_left",
                                                ifelse(df_complete$Shift_direction_ctrl_to_rnase == "right shift", "RBP_right", "RBP")),
                                         "Non-RBP")


# grid for regressionplane
pI_seq <- seq(min(df_complete$pI), max(df_complete$pI), length.out = 30)
Mass_seq <- seq(min(df_complete$Mass_kDa), max(df_complete$Mass_kDa), length.out = 30)
grid <- expand.grid(pI = pI_seq, Mass_kDa = Mass_seq)
grid$predicted <- predict(model, newdata = grid)

# determine colors
farben <- list("RBP_left" = "#c9a0dc",     # hell-lila
               "RBP_right" = "#7a42a6",    # dunkel-lila
               "RBP" = "purple",
               "Non-RBP" = "lightgray")

# Plotly Plot 
fig <- plot_ly()

# RBP_left
fig <- fig %>%
  add_markers(data = subset(df_complete, RBP_shift_category == "RBP_left"),
              x = ~pI, y = ~Mass_kDa, z = ~probability,
              marker = list(color = farben[["RBP_left"]], size = 4),
              name = "RBP (left shift)")

# RBP_right
fig <- fig %>%
  add_markers(data = subset(df_complete, RBP_shift_category == "RBP_right"),
              x = ~pI, y = ~Mass_kDa, z = ~probability,
              marker = list(color = farben[["RBP_right"]], size = 4),
              name = "RBP (right shift)")

# RBP without Shift-Direction 
fig <- fig %>%
  add_markers(data = subset(df_complete, RBP_shift_category == "RBP"),
              x = ~pI, y = ~Mass_kDa, z = ~probability,
              marker = list(color = farben[["RBP"]], size = 4),
              name = "RBP (no shift info)")

# Non-RBP
fig <- fig %>%
  add_markers(data = subset(df_complete, RBP_shift_category == "Non-RBP"),
              x = ~pI, y = ~Mass_kDa, z = ~probability,
              marker = list(color = farben[["Non-RBP"]], size = 4),
              name = "Non-RBP")

# Regressionplane
fig <- fig %>%
  add_surface(x = ~pI_seq,
              y = ~Mass_seq,
              z = matrix(grid$predicted, nrow = 30),
              opacity = 0.5,
              colorscale = "Purples",
              showscale = FALSE,
              name = "Regression Surface")

# Layout
fig <- fig %>%
  layout(scene = list(
    xaxis = list(title = "pI"),
    yaxis = list(title = "Mass (kDa)", tick0 = 0, dtick = 100, range = c(0, 1000)),
    zaxis = list(title = "Probability")
  ),
  legend = list(title = list(text = "<b>Protein Classification</b>")))

fig


```
## check individual relation between predictors and probability
### determine whether there is a signifcant relation between a higher pI and an increased probability for RNA-binding
- using a t-test the pIs of RBPs and Non-RBPs are compared
```{r}

#hypothesis: proteins with a higher pI are more significantly more likely to be RBPs -> have positively charged domains at physiological pH -> are more likely to bind negatively charged RNA

# Group proteins based on optimal threshold:
df_complete$RBP_group <- ifelse(df_complete$probability > optimal_threshold,
                               "RBP", "Non-RBP")

# Optional: check mean pI values per group to confirm
aggregate(pI ~ RBP_group, data = df_complete, mean)

# Perform one-sided Welch Two Sample t-test comparing pI of RBP vs Non-RBP
# as big sample sizes are used, it was assumed that the central limit theorem applies (meaning the data follows a normal distribution)
t_test_result <- t.test(
  df_complete$pI[df_complete$RBP_group == "RBP"],
  df_complete$pI[df_complete$RBP_group == "Non-RBP"],
  alternative = "greater"
)

print(t_test_result)

```
output:
- there is a signifcant difference between the pIs of RBPs and Non-RBPs
-> RBPs tend to have a higher pI

### visualization of relation between pI and probability
```{r}


# create boxplot
boxplot(pI ~ RBP_group,
        data = df_complete,
        col = c("#A9C1A6", "#c9a0dc"),
        main = "compare pI\nbetween RBP and Non-RBP",
        xlab = "group",
        ylab = "isoelectric point (pI)")

# calculate mean
mean_rbp <- mean(df_complete$pI[df_complete$RBP_group == "RBP"])
mean_nonrbp <- mean(df_complete$pI[df_complete$RBP_group == "Non-RBP"])

print(mean_rbp)
print(mean_nonrbp)

# depict mean in plot
points(c(1, 2), c(mean_nonrbp, mean_rbp), pch = 19, col = "coral")

# Oadd legend
legend("topright", legend = c("Mean"), col = "coral", pch = 19)

```
### determine whether there is a signifcant relation between mass and the probability for RNA-binding
- using a t-test the mass of RBPs and Non-RBPs are compared
```{r}
#check whether mass differs significantly between RBPs and Non-RBPs

# Perform two-sided Welch Two Sample t-test comparing pI of RBP vs Non-RBP
t_test_result <- t.test(
  df_complete$Mass_kDa[df_complete$RBP_group == "RBP"],
  df_complete$Mass_kDa[df_complete$RBP_group == "Non-RBP"],
)

print(t_test_result)
```
output:
- there is no significant relation between mass and probability
-> the signficant relation only exists after accounting for pI in the context of the multiple regression model



# 5. dimension reduction

### PCA
Goal: Reduce dimension and further visualize data in clusters with only the most important structures. Batch cleaned data is used in order to ensure no technical variation. 
##Control

#check how data set is made up, ctrl
```{r}
ctrl_perc_df <- as.data.frame(percent_ctrl_df[,-1])
rownames(ctrl_perc_df) <- ctrl_perc_df$X
C_test <- ctrl_perc_df
str(C_test)

# we have 3680 observations and 75 features 
```

#Test data
-> check if varibales are on the same scale 
```{r}
apply(C_test, 2, mean) 
```
-> mean is close to 0, centered enough
```{r}
apply(C_test, 2, sd)
```
-> big differences, not perfectly scaled 

##pca selbst ctrl 
-> prcomp: uses singular value decomposition, according to r it has slightly better numerical accuracy
-> center=TRUE, scale=TRUE: makes sure mean ~ 0 , sd ~ 1 -> Z-scores 
```{r}
ctrl_pca <- prcomp(ctrl_perc_df, center = TRUE, scale = TRUE)

names(ctrl_pca)
```

```{r}
head(ctrl_pca$x, 10)
```

-> eigenvalues stored in "rotation" 
```{r}
ctrl_pca$rotation
```

-> standard deviation of each principal component is saved in "sdev", as a measure of variance
-> check how much do points vary on average from the mean
```{r}
ctrl_pca$sdev

```

-> get variance by squaring the standard deviation
-> check how much does the data spread out in square units 
```{r}
ctrl_pca_var <- ctrl_pca$sdev^2 # eigenvalues 
ctrl_pca_var

```

Goal: 
-> control which PC explains how much variance 
```{r}
ctrl_pca_ve <- ctrl_pca_var / sum(ctrl_pca_var) 
ctrl_pca_ve_percent <- ctrl_pca_ve * 100
ctrl_pca_ve_percent
```
-> PC1: 40.9%, PC2:22.6%, PC3:13.47%



#prozentplot ctrl

```{r}
#1. plot
plot(ctrl_pca_ve_percent, type = "b", pch = 19,
     xlab = "Principal Component",
     ylab = "Explained Variance (%)",
     main = "Scree Plot: Explained Variance per PC, ctrl")
```


#barplot ctrl

```{r}
#plot 
bar_positions <- barplot(ctrl_pca_ve_percent,
                         main = "Explained Variance per PC (in %), ctrl",
                         xlab = "Principal Component",
                         ylab = "Explained Variance (%)",
                         col = "#c9a0dc",
                        las = 2,
                         xaxt = "n")  # keine x-Achsen-Beschriftung

axis(1, at = bar_positions, labels = paste0("PC", 1:length(ctrl_pca_ve_percent)), las = 2)



```


##create biplot 
###plot 
```{r}
#1. plot
biplot(ctrl_pca, scale = 0, cex = 0.6,
       col = c("black", "#c9a0dc"),
       main="PCA biplot, ctrl",
       xlab = "PC1 (44% variance explained)",  
       ylab = "PC2 (21% variance explained)")
```

#test graph 
```{r}

library(ggbiplot)
#1. axis
my_var_explained <- c(40.95, 22.63)  
#2. plot
ggbiplot(ctrl_pca,
         obs.scale = 1,
         var.scale = 1,
         ellipse = FALSE,
         circle = TRUE,
         var.axes = TRUE) +
  xlab(paste0("PC1 (", round(my_var_explained[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(my_var_explained[2], 1), "%)")) +
  theme_minimal()+
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 14)) +
  guides(colour = "none") 
   


```



###determine cluster
#elbowplot 
Goal: determine what amount of cluster is suitable for kmeans clustering 
```{r} 
#1. calculate wss: within-cluster sum of squares
wss <- sapply(1:25, function(k){
  kmeans(ctrl_pca$x, centers = k, nstart = 25)$tot.withinss
})

#2. plot
plot(1:25, wss, type = "b", pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Total within-cluster-variance",
     main = "Elbow method to determin optimal cluster amount, ctrl") 
```
#Interpretation: 
-> no clear bend, potentially at 2 
#silhouette score 
Goal: determine what amount of cluster is suitable for kmeans clustering 

```{r}
# Silhouette analysis 

library(cluster)
library(factoextra)

# 1. Extract PCA-transformed data
ctrl_sil_df <- ctrl_pca$x

# 2. Placeholder for average silhouette widths
ctrl_avg_sil <- numeric(length = 9)  # for k = 2 to 10

# 3. Calculate silhouette scores for different k values
k_values <- 2:10
for (i in seq_along(k_values)) {
  k <- k_values[i]
  ctrl_km <- kmeans(ctrl_sil_df, centers = k, nstart = 25)
  ctrlc_ss <- silhouette(ctrl_km$cluster, dist(ctrl_sil_df))
  ctrl_avg_sil[i] <- mean(ctrlc_ss[, 3])
}

# 4. Plot similar to elbow method
plot(k_values, ctrl_avg_sil, type = "b", pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Average silhouette width",
     main = "Silhouette analysis to determine the optimum number of clusters, ctrl")

# 5. Identify best k based on max silhouette width
best_k <- k_values[which.max(ctrl_avg_sil)]

# 6. Run KMeans and compute silhouette for best k
km_best <- kmeans(ctrl_sil_df, centers = best_k, nstart = 25)
ctrl_best_sil <- silhouette(km_best$cluster, dist(ctrl_sil_df))

# 7. Visualize silhouette plot
fviz_silhouette(
  ctrl_best_sil,
  palette = c("#c9a0dc", "#A9C1A6"),
  ggtheme = theme_classic(),
  main = paste("Ctrl silhouette plot for k =", best_k)
)

```
#Interpretation
k=2 seems most suitable

###kmeans 
Goal:divide dataset into k cluster to understand grouping and potentionally find biological relations 


```{r}
#1. apply kmeans to reduced data set 
# using k=2
#kmeans runs 25 and uses best result
C_kmeans_result <- kmeans(ctrl_pca$x, centers = 2, nstart = 25)
```


```{r}
#1. in which cluster every obserbation is assigned to 
C_kmeans_result$cluster    

#2. coordinates the cluster center
C_kmeans_result$centers     

#3. shows how compact the clusters are using WSS
# lower WSS means a more compact cluster
C_kmeans_result$tot.withinss
```


```{r}
#1. prepare data
#save pca matrix into data frame 
# add cluster column
ctrl_data <- as.data.frame(ctrl_pca$x)
ctrl_data$cluster <- as.factor(C_kmeans_result$cluster)

#2. plot
library(ggplot2)

ggplot(ctrl_data, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "K-means clusters in PCA area, ctrl",
       x = "Principal component 1 (40.95%)",
       y = "Principal component 2 (22.63%)") +
  theme_minimal() +
  scale_color_manual(values = c("#c9a0dc","#A9C1A6"))
```

#plot PC1 and PC2 as control 
```{r}
first_two_components <- ctrl_data[,1:2]
plot(first_two_components)
```

###PCA RNASE

##Check how dataset is made up  

```{r}
rnase_perc_df <- as.data.frame(percent_rnase_df[,-1])
rownames(rnase_perc_df) <- rnase_perc_df$X
R_TEST <- rnase_perc_df
str(R_TEST)
# we have 3680 observations and 75 features 
```

```{r}
head(rnase_perc_df)
```

#Test data
-> check if varibales are on the same scale 

```{r}
apply(R_TEST, 2, mean) 
```
-> mean is close to 0, centered enough


```{r}
apply(R_TEST, 2, sd)
```
-> big differences, not perfectly scaled 

###pca selbst 
-> prcomp: uses singular value decomposition, according to r it has slightly better numerical accuracy
-> center=TRUE, scale=TRUE: makes sure mean ~ 0 , sd ~ 1 -> Z-scores

```{r}
rnase_pca <- prcomp(rnase_perc_df, center = TRUE, scale = TRUE)

names(rnase_pca)
```



-> eigenvalues stored in "rotation" 
```{r}
rnase_pca$rotation
```


-> The standard deviation of each principal component is saved in "sdev", as a measure of variance
-> check how much points vary on average from the mean
```{r}
rnase_pca$sdev

```

-> get variance by squaring the standard deviation
-> check how much the data spreads out in square units 
```{r}
rnase_pca_var <- rnase_pca$sdev^2 # eigenvalues 
rnase_pca_var

```


Goal: 
-> control which PC explains how much variance 
```{r}
rnase_pca_ve <- rnase_pca_var / sum(rnase_pca_var) 
rnase_pca_ve_percent <- rnase_pca_ve * 100
rnase_pca_ve_percent

```
-> PC1: 42.05%, PC2:19.35%, PC3:15.14 %


#prozentplot 

```{r}
#1. plot
plot(rnase_pca_ve_percent, type = "b", pch = 19,
     xlab = "Principal Component",
     ylab = "Explained Variance (%)",
     main = "Scree Plot: Explained Variance per PC, RNAse")

```

#barplot
```{r}
#plot 
bar_positions <- barplot(rnase_pca_ve_percent,
                         main = "Explained Variance per PC (in %), RNAse",
                         xlab = "Principal Component",
                         ylab = "Explained Variance (%)",
                         col = "#c9a0dc",
                        las = 2,
                         xaxt = "n")  

axis(1, at = bar_positions, labels = paste0("PC", 1:length(rnase_pca_ve_percent)), las = 2)

```

#biplot
```{r}
#1. plot
biplot(rnase_pca, scale = 0, cex = 0.6,
       col = c("black", "#c9a0dc"),                
       xlab = "PC1 (42.05% variance explained)",  
       ylab = "PC2 (19.35% variance explained)", 
       main="PCA biplot, RNAse")

```

#test graph 
```{r}
library(ggbiplot)
#1. axis
my_var_explained <- c(42.05, 19.35)  
#2. plot
ggbiplot(rnase_pca,
         obs.scale = 1,
         var.scale = 1,
         ellipse = FALSE,
         circle = TRUE,
         var.axes = TRUE) +
  xlab(paste0("PC1 (", round(my_var_explained[1], 1), "%)")) +
  ylab(paste0("PC2 (", round(my_var_explained[2], 1), "%)")) +
  theme_minimal()

```



###determine cluster
#elbowplot 
Goal: determine what amount of clusters is suitable for K-means clustering 

```{r} 
#1. calculate wss: within-cluster sum of squares 
wss <- sapply(1:25, function(k){
  kmeans(rnase_pca$x, centers = k, nstart = 25)$tot.withinss
})
#2. plot
plot(1:25, wss, type = "b", pch = 19,
     xlab = "Number of clusters (k)",
     ylab = "Total within-cluster-variance",
     main = "Elbow method to determin optimal cluster amount, RNAse") 
```
#Interpretation: 
-> no clear bend, potentially at 2 


#silhouette score 
Goal: determine what number of clusters is suitable for K-means clustering 

```{r}
# Silhouette analysis 

library(cluster)
library(factoextra)

#1. Extract PCA-transformed data
rnase_sil_df <- rnase_pca$x

#2. Placeholder for average silhouette widths
rnase_avg_sil <- numeric(length = 9)  

# 3. Calculate silhouette scores for different k values
k_values <- 2:10
for (i in seq_along(k_values)) {
  k <- k_values[i]
  rnase_km <- kmeans(rnase_sil_df, centers = k, nstart = 25)
  rnase_ss <- silhouette(rnase_km$cluster, dist(rnase_sil_df))
  rnase_avg_sil[i] <- mean(rnase_ss[, 3])
}

# 4. Plot similar to elbow method
plot(k_values, rnase_avg_sil, type = "b", pch = 19,
     xlab = "Number of clusters (k)",
     ylab =  "Average silhouette width",
     main = "Silhouette analysis to determine the optimum number of clusters, RNAse")

# 5. Identify the best k based on max silhouette width
best_k <- k_values[which.max(rnase_avg_sil)]

# 6. Run K-Means and compute the  silhouette for the best k
rnase_km_best <- kmeans(rnase_sil_df, centers = best_k, nstart = 25)
rnase_best_sil <- silhouette(rnase_km_best$cluster, dist(rnase_sil_df))

# 7. Visualize silhouette plot
palette_colors <- c("#c9a0dc", "#A9C1A6", "#F5E1A4")[1:best_k]

fviz_silhouette(
  rnase_best_sil,
  palette = palette_colors,
  ggtheme = theme_classic(),
  main = paste("RNAse silhouette plot for k =", best_k)
)

```
#Interpretation
k=2 seems most suitable


###kmeans 
Goal:divide dataset into k cluster to understand grouping and potentionally find biological relations 

#apply K-means rnase
```{r}
#1. apply kmeans to  data set 
# using k=2
#kmeans runs 25 and uses best result
R_kmeans_result <- kmeans(rnase_pca$x, centers = 2, nstart = 25)
```

```{r}
#1. in which cluster every obserbation is assigned to 
R_kmeans_result$cluster    

#2. Coordinates the cluster center
R_kmeans_result$centers   

#3. shows how compact the clusters are using WSS
# lower WSS means a more compact cluster
R_kmeans_result$tot.withinss
```

#K-means Plot rnase
```{r}
#1. prepare data
#save pca matrix into data frame 
# add cluster column
rnase_data <- as.data.frame(rnase_pca$x)
rnase_data$cluster <- as.factor(R_kmeans_result$cluster)

#2. plot
library(ggplot2)

ggplot(rnase_data, aes(x = PC1, y = PC2, color = cluster)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "K-means clusters in PCA area, RNAse",
       x = "Principal component 1 (42.05%)",
       y = "Principal component 2 (19.35%)") +
  theme_minimal() +
  scale_color_manual(values = c("#c9a0dc","#A9C1A6"))
```

#plot PC1 and PC2 as control
```{r}
first_two_components <- rnase_data[,1:2]
plot(first_two_components)
```


###right shift
```{r}
#create dataframe with right shift proteins only
# convert matrices to dataframes
corrected_ctrl_df <- as.data.frame(S_corrected_expr_ctrl, stringsAsFactors = FALSE)
corrected_rnase_df <- as.data.frame(S_corrected_expr_rnase, stringsAsFactors = FALSE)

# add column with protein names to newly created dataframes
corrected_ctrl_with_protein <- cbind(Protein = protein_names, corrected_ctrl_df)
corrected_rnase_with_protein <- cbind(Protein = protein_names, corrected_rnase_df)

# filter for matching columns between right_shift_proteins_df and the batch removed dataframes
right_shift_ctrl <- corrected_ctrl_df[corrected_ctrl_with_protein$Protein %in% right_shift_proteins_df$Protein, ]
right_shift_rnase <- corrected_rnase_df[corrected_rnase_with_protein$Protein %in% right_shift_proteins_df$Protein, ]

#depict result
head(right_shift_ctrl)
head(right_shift_rnase)

```


### label rsp in larger pca 
Goal: highlight right shifts proteins in a visualization of PC1 and PC2 to determine biological relation

##rnase labeling

```{r}
library(ggplot2)

#1. prepare PCA, extract numerical values
test_rnase_df <- as.data.frame(percent_rnase_df[,-1])
TR_pca_result <- prcomp(test_rnase_df, center = TRUE, scale. = TRUE)

#2. convert PCA results into dataframe
TR_pca_df <- as.data.frame(TR_pca_result$x)

#3. copy protein names from rownames
TR_pca_df$Protein <- percent_rnase_df$Protein

#4. define righ shift proteins
test_highlight_proteins <- right_shift_proteins_df$Protein

#5. add column with highlights
TR_pca_df$Highlight <- ifelse(TR_pca_df$Protein %in% test_highlight_proteins, "Yes", "No")
unique(TR_pca_df$Highlight)

#6. show how many right shift proteins were detected
table(TR_pca_df$Highlight)

#7. plot
ggplot(TR_pca_df, aes(x = PC1, y = PC2, color = Highlight)) +
  geom_point(alpha = 0.8, size = 2) +
  scale_color_manual(
  values = c("Yes" = "#c9a0dc", "No" = "lightgrey"),
  name = "Highlighted Proteins",
  labels = c("Yes" = "Right shift proteins", "No" = "Other proteins")
) +
  labs(title = "PCA Plot with Highlighted Proteins",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal()+
  theme(
    legend.position = "right",       
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 10)
  )

```


```{r}
head(right_shift_rnase)
```


##ctrl

```{r}
library(ggplot2)

#1. prepare PCA, extract numerical values
test_ctrl_df <- as.data.frame(percent_ctrl_df[,-1])
TC_pca_result <- prcomp(test_ctrl_df, center = TRUE, scale. = TRUE)

#2. convert PCA results into dataframe
TC_pca_df <- as.data.frame(TC_pca_result$x)

#3. copy protein names from rownames
TC_pca_df$Protein <- percent_ctrl_df$Protein

#4. define righ shift proteins
test_highlight_proteins <- right_shift_proteins_df$Protein

#5. add column with highlights
TC_pca_df$Highlight <- ifelse(TC_pca_df$Protein %in% test_highlight_proteins, "Yes", "No")
unique(TC_pca_df$Highlight)

#6. show how many right shift proteins were detected
table(TC_pca_df$Highlight)

#7. plot
ggplot(TC_pca_df, aes(x = PC1, y = PC2, color = Highlight)) +
  geom_point(alpha = 0.8, size = 2) +
  scale_color_manual(
  values = c("Yes" = "#c9a0dc", "No" = "lightgrey"),
  name = "Highlighted Proteins",
  labels = c("Yes" = "Right shift proteins", "No" = "Other proteins")
) +
  labs(title = "PCA Plot with Highlighted Proteins",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal()+
  theme(
    legend.position = "right",       
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 10)
  )

```





##label RNA binding proteins in K-means

#isolate rbp 

```{r}
# Select only relevant columns from the RBP and Non-RBP tables
rbp_subset <- Table_HS_RBP[, c("Entry_Name", "pI", "Mass_kDa" , "Length_AA")]
non_rbp_subset <- Table_HS_Non_RBP[, c("Entry_Name", "pI", "Mass_kDa", "Length_AA")]

# Combine both subsets into a single table
Table_HS_Combined <- rbind(rbp_subset, non_rbp_subset)

# # 3. Merge: Proteinname-Spalten heißen unterschiedlich -> by.x / by.y verwenden
merged_df <- merge(test_scores_combined_df, Table_HS_Combined,
                   by.x = "Protein", by.y = "Entry_Name",
                   all.x = TRUE)  # keep all entries from test_scores_combined_df
#depict result
head(merged_df)
dim(merged_df)
```

```{r}
#1.filter for RBPs
merged_df <- merge(test_scores_combined_df, Table_HS_Combined,
                   by.x = "Protein", by.y = "Entry_Name",
                   all.x = TRUE)
df_complete <- merged_df[complete.cases(merged_df[, c("probability", "pI", "Mass_kDa")]), ]
df_complete$RBP_shift_category <- ifelse(df_complete$probability >= optimal_threshold,
                                         ifelse(df_complete$Shift_direction_ctrl_to_rnase == "left shift", "RBP_left",
                                                ifelse(df_complete$Shift_direction_ctrl_to_rnase == "right shift", "RBP_right", "RBP")),
                                         "Non-RBP")

rbp_only <- df_complete[df_complete$RBP_shift_category %in% c("RBP_left", "RBP_right"), ]

#2.extract RBPS and store in a separate vector
rbp_proteins <- rbp_only$Protein
head(rbp_proteins)

```




#K-means ctrl 
```{r}
#1. convert PCA ctrl data into dataframe 
Ctrl_data <- as.data.frame(ctrl_pca$x)
Ctrl_data$cluster <- as.factor(C_kmeans_result$cluster)

#2. isolate protein names
Ctrl_data$protein_name <- c(percent_ctrl_df$Protein)
#check if rbp present 
Ctrl_data$is_rbp <- Ctrl_data$protein_name %in% rbp_proteins


#3.plot
library(ggplot2)

ggplot(Ctrl_data, aes(x = PC1, y = PC2, color = cluster, shape = is_rbp)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "RBP label in K-means, ctrl",
       x = "Hauptkomponente 1 (PC1)",
       y = "Hauptkomponente 2 (PC2)",
       color = "Cluster",
       shape = "Protein-Typ") +
  theme_minimal() +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17)) + 
  scale_color_manual(values = c("1" = "#c9a0dc", "2" = "lightgrey"))


ggplot(Ctrl_data,aes(x = PC1, y = PC2, color = cluster)) +
  
  scale_shape_manual(values = c("non-RBP" = 16, "RBP" = 17)) +
  labs(shape = "Protein-Typ")

```


#K-means rnase 
```{r}
#1. convert PCA ctrl data into dataframe 
rnase_data <- as.data.frame(rnase_pca$x)
rnase_data$cluster <- as.factor(R_kmeans_result$cluster)

#2. isolate protein names
rnase_data$protein_name <- c(S_corrected_expr_rnase_df$Protein)
#check if rbp present 
rnase_data$is_rbp <- rnase_data$protein_name %in% rbp_proteins


#3.plot
library(ggplot2)

ggplot(rnase_data, aes(x = PC1, y = PC2, color = cluster, shape = is_rbp)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "RBP label in K-means, RNAse",
       x = "Hauptkomponente 1 (PC1)",
       y = "Hauptkomponente 2 (PC2)",
       color = "Cluster",
       shape = "Protein-Typ") +
  theme_minimal() +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17)) + 
  scale_color_manual(values = c("1" = "#c9a0dc", "2" = "lightgrey"))

ggplot(Ctrl_data,aes(x = PC1, y = PC2, color = cluster)) +
  
  scale_shape_manual(values = c("non-RBP" = 16, "RBP" = 17)) +
  labs(shape = "Protein-Typ")

```



###CHI squared test 

Goal: Determine if cluster are associated with RNA binding proteins 

#rnase

```{r}
#1. Add protein column
rnase_data$protein_name <-  c(percent_rnase_df$Protein)

#2.set TRUE to RBP and FALSE to non - RBP
rnase_data$is_rbp <- rnase_data$protein_name %in% rbp_proteins
#3.turn cluster into a factor 
rnase_data$cluster <- factor(rnase_data$cluster, levels = c(1, 2))  
#keep cluster as order
```



```{r}
#1. create table which sorts RBPs and non-RBPs into clusters
r_tbl <- table(rnase_data$is_rbp, rnase_data$cluster)
show(r_tbl)
```

```{r}
#1.table counts the amount of RBPs
table(rnase_data$is_rbp)
str(rnase_data$is_rbp)
```

```{r}
print(rnase_data)
```


```{r}
#1. apply chi sqquared to earlier created table
chisq.test(r_tbl)


```
#Interpretation:
p=4.237e-09 < 0.05 -> significant

##ctrl
```{r}
#1. Add protein column
Ctrl_data$protein_name <-  c(percent_ctrl_df$Protein)
#2.set TRUE to RBP and FALSE to non - RBP
Ctrl_data$is_rbp <- Ctrl_data$protein_name %in% rbp_proteins
#3.turn cluster into a factor 
Ctrl_data$cluster <- factor(Ctrl_data$cluster, levels = c(1, 2))  
#keep cluster as order
```

```{r}
#1. create table which sorts RBPs and non-RBPs into clusters
c_tbl <- table(Ctrl_data$is_rbp, Ctrl_data$cluster)
show(c_tbl)
```

```{r}
#1.table counts the amount of RBPs
table(Ctrl_data$is_rbp)
str(Ctrl_data$is_rbp)
```


```{r}
print(ctrl_data)
```


```{r}
#1. apply chi sqquared to earlier created table
chisq.test(c_tbl)


```
#Interpretation'
p=2.2e-16 < 0.05 -> significant: there is a connection between clusters found through PCA and classification of proteins as RBPs





```{r}
#1.table counts the amount of RBPs
table(test_ctrl_data$is_rbp)
str(test_ctrl_data$is_rbp)
```


#label RBP in K-means 
#rnase
```{r}
#1. convert PCA ctrl data into dataframe 
test_rnase_data <- as.data.frame(test_rnase_pca$x)
test_rnase_data$cluster <- as.factor(TR_kmeans_result$cluster)

#2. isolate protein names
test_rnase_data$protein_name <- c(percent_rnase_df$Protein)
#check if rbp present 
test_rnase_data$is_rbp <- test_rnase_data$protein_name %in% rbp_proteins


#3.plot
library(ggplot2)

ggplot(test_rnase_data, aes(x = PC1, y = PC2, color = cluster, shape = is_rbp)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "K-Means Cluster im PCA-Raum",
       x = "Hauptkomponente 1 (PC1)",
       y = "Hauptkomponente 2 (PC2)",
       color = "Cluster",
       shape = "Protein-Typ") +
  theme_minimal() +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17)) + 
  scale_color_manual(values = c("1" = "#c9a0dc", "2" = "lightgrey"))

ggplot(test_rnase_data,aes(x = PC1, y = PC2, color = cluster)) +
  
  scale_shape_manual(values = c("non-RBP" = 16, "RBP" = 17)) +
  labs(shape = "Protein-Typ")

```


#ctrl

```{r}
#1. convert PCA ctrl data into dataframe 
test_ctrl_data <- as.data.frame(test_ctrl_pca$x)
test_ctrl_data$cluster <- as.factor(TC_kmeans_result$cluster)

#2. isolate protein names
test_ctrl_data$protein_name <- c(percent_ctrl_df$Protein)
#check if rbp present 
test_ctrl_data$is_rbp <- test_ctrl_data$protein_name %in% rbp_proteins


#3.plot
library(ggplot2)

ggplot(test_ctrl_data, aes(x = PC1, y = PC2, color = cluster, shape = is_rbp)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "K-Means Cluster im PCA-Raum",
       x = "Hauptkomponente 1 (PC1)",
       y = "Hauptkomponente 2 (PC2)",
       color = "Cluster",
       shape = "Protein-Typ") +
  theme_minimal() +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 17)) + 
  scale_color_manual(values = c("1" = "#c9a0dc", "2" = "lightgrey"))

ggplot(test_ctrl_data,aes(x = PC1, y = PC2, color = cluster)) +
  
  scale_shape_manual(values = c("non-RBP" = 16, "RBP" = 17)) +
  labs(shape = "Protein-Typ")

```


##RSP in PCA 
#rnase labeling

```{r}
library(ggplot2)

#1. prepare PCA, extract numerical values
test_rnase_df <- as.data.frame(percent_rnase_df[,-1])
TR_pca_result <- prcomp(test_rnase_df, center = TRUE, scale. = TRUE)

#2. convert PCA results into dataframe
TR_pca_df <- as.data.frame(TR_pca_result$x)

#3. copy protein names from rownames
TR_pca_df$Protein <- percent_rnase_df$Protein

#4. define righ shift proteins
test_highlight_proteins <- right_shift_proteins_df$Protein

#5. add column with highlights
TR_pca_df$Highlight <- ifelse(TR_pca_df$Protein %in% test_highlight_proteins, "Yes", "No")
unique(TR_pca_df$Highlight)

#6. show how many right shift proteins were detected
table(TR_pca_df$Highlight)

#7. plot
ggplot(TR_pca_df, aes(x = PC1, y = PC2, color = Highlight)) +
  geom_point(alpha = 0.8, size = 2) +
  scale_color_manual(
  values = c("Yes" = "#c9a0dc", "No" = "lightgrey"),
  name = "Highlighted Proteins",
  labels = c("Yes" = "Right shift proteins", "No" = "Other proteins")
) +
  labs(title = "PCA Plot RNAse with Highlighted Proteins",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal()+
  theme(
    legend.position = "right",       
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 10)
  )

```


#ctrl

```{r}
library(ggplot2)

#1. prepare PCA, extract numerical values
test_ctrl_df <- as.data.frame(percent_ctrl_df[,-1])
TC_pca_result <- prcomp(test_ctrl_df, center = TRUE, scale. = TRUE)

#2. convert PCA results into dataframe
TC_pca_df <- as.data.frame(TC_pca_result$x)

#3. copy protein names from rownames
TC_pca_df$Protein <- percent_ctrl_df$Protein

#4. define righ shift proteins
C_test_highlight_proteins <- right_shift_proteins_df$Protein
print(C_test_highlight_proteins)
class(C_test_highlight_proteins)

#5. add column with highlights
TC_pca_df$Highlight <- ifelse(TC_pca_df$Protein %in% C_test_highlight_proteins, "Yes", "No")
unique(TC_pca_df$Highlight)
print(TC_pca_df$Protein)
class(test_highlight_proteins)
#6. show how many right shift proteins were detected
table(TC_pca_df$Highlight)

#7. plot
ggplot(TC_pca_df, aes(x = PC1, y = PC2, color = Highlight)) +
  geom_point(alpha = 0.8, size = 2) +
  scale_color_manual(
  values = c("Yes" = "#c9a0dc", "No" = "lightgrey"),
  name = "Highlighted Proteins",
  labels = c("Yes" = "Right shift proteins", "No" = "Other proteins")
) +
  labs(title = "PCA Plot Ctrl with Highlighted Proteins",
       x = "Principal Component 1",
       y = "Principal Component 2") +
  theme_minimal()+
  theme(
    legend.position = "right",       
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 10)
  )

```


